
\renewcommand{\rwpre}{./graph/DMDK}
\section{Time-splitting, time-scales and SMR}
This exercise is about mortaity among Danish Diabetes patients. It is
based on the dataset \texttt{DMlate}, a random sample of 10,000
patients from the Danish Diabetes Register (scrambeled dates), all
with date of diagnosis after 1994.
\begin{enumerate}
\item First load the data and take a look at the data:
\begin{Schunk}
\begin{Sinput}
 library( Epi)
 library(popEpi)
 library( mgcv)
 library(tidyverse)
 sessionInfo()
 data( DMlate)
 str( DMlate)
\end{Sinput}
\end{Schunk}
You can get a more detailed explanation of the data by referring to
the help page:
\begin{Schunk}
\begin{Sinput}
 ?DMlate
\end{Sinput}
\end{Schunk}
\item Set up the dataset as a \texttt{Lexis} object with age, calendar
  time and duration of diabetes as timescales, and date of death as
  event. Make sure that you know what each of the arguments to
  \texttt{Lexis} mean:
\begin{Schunk}
\begin{Sinput}
 LL <- Lexis(entry = list(A = dodm-dobth,
                          P = dodm,
                        dur = 0),
              exit = list(P = dox),
       exit.status = factor(!is.na(dodth),
                            labels = c("Alive","Dead")),
              data = DMlate)
\end{Sinput}
\end{Schunk}
Take a look at the first few lines of the resulting dataset, for
example using \texttt{head()}.
\item Get an overview of the mortality by using \texttt{stat.table}
  to tabulate no. deaths, person-years (\texttt{lex.dur}) and the
  crude mortality rate by sex. Try:
\begin{Schunk}
\begin{Sinput}
 stat.table(sex,
            list(D = sum(lex.Xst == "Dead"),
                 Y = sum(lex.dur),
              rate = ratio(lex.Xst == "Dead", lex.dur, 1000)),
            margins = TRUE,
            data = LL)
\end{Sinput}
\end{Schunk}
\item If we want to assess how mortality depends on age, calendar time
  and duration or how it relates to population mortality, we should
  in principle split the follow-up along all
  three time scales. In practice it is sufficient to split it along
  one of the time-scales and then use the value of each of the
  time-scales at the left endpoint of the intervals.
  Use \texttt{splitLexis} (or \texttt{splitMulti} from the
  \texttt{popEpi} package) to split the follow-up along the
  age-axis in sutiable intervals (here set to 1/2 year, but really
  immaterial as long as it is small):
\begin{Schunk}
\begin{Sinput}
 SL <- splitLexis(LL, breaks = seq(0,125,1/2), time.scale = "A")
 summary(SL)
\end{Sinput}
\end{Schunk}
  How many records are now in the dataset? How many person-years?
  Compare to the original \texttt{Lexis}-dataset.
\end{enumerate}
\subsection*{SMR}
The SMR is the \textbf{S}tandardized \textbf{M}ortality
\textbf{R}atio, which is the mortality rate-ratio between the diabetes
patients and the general population.  In real studies we would
subtract the deaths and the person-years among the diabetes patients
from those of the general population, but since we do not have access
to these, we make the comparison to the general population at large,
\textit{i.e.} also including the diabetes patients.
So we now want to include the population mortality rates as a fixed
variable in the split dataset; for each record in the split dataset we
attach the value of the population mortality for the relevant sex, and
and calendar time.
This can be achieved in two ways: Either we just use the current split
of follow-up time and allocate the population mortality rates for some
suitably chosen (mid-)point of the follow-up in each, or we make a
second split by date, so that follow-up in the diabetes patients is in
the same classification of age and data as the population mortality
table.
\begin{enumerate}[resume]
\item We will use the former approach, using the dataset split in
  6 month intervals, and then include as an extra variable the
  population mortality as available from the data set
  \texttt{M.dk}.
  First create the variables in the diabetes dataset that we need
  for matching with the population mortality data, that is sex and
  age and date at the midpoint of each of the intervals (or rater at a
  point 3 months after the left endpoint of the interval --- recall
  we split the follow-up in 6 month intervals).
  We need to have variables of the same type when we merge, so we must
  transform the sex variable in \texttt{M.dk} to a factor, and must
  for each follow-up interval in the \texttt{SL} data have an age and
  a period variable that can be used in merging with the population data.
\begin{Schunk}
\begin{Sinput}
 str(SL)
 SL$Am <- floor(SL$A + 0.25)
 SL$Pm <- floor(SL$P + 0.25)
 data(M.dk)
 str(M.dk)
 M.dk <- transform(M.dk, Am = A,
                         Pm = P,
                        sex = factor(sex, labels = c("M","F")))
 str(M.dk)
\end{Sinput}
\end{Schunk}
Then match the rates from \texttt{M.dk} into \texttt{SL} ---
\texttt{sex}, \texttt{Am} and \texttt{Pm} are the common variables,
and therefore the match is on these variables:
\begin{Schunk}
\begin{Sinput}
 SLr <- merge(SL, M.dk[,c("sex", "Am", "Pm", "rate")])
 dim(SL)
 dim(SLr)
\end{Sinput}
\end{Schunk}
This merge (remember to \texttt{?merge}!) only takes rows that have
information from both datasets, hence the slightly fewer rows in
\texttt{SLr} than in \texttt{SL}.
\item Compute the expected number of deaths as the person-time
   multiplied by the corresponding population rate, and put it in a
   new variable, \texttt{E}, say (\texttt{E}xpected). Use \texttt{stat.table}
   to make a table of observed, expected and the ratio (SMR) by age
   (suitably grouped, look for \texttt{cut}) and sex.
\item Fit a poisson model with sex as the explanatory varable and
  log-expected as offset to derive the SMR (and c.i.).
  Some of the population mortality rates are 0, so you need to exclude
  those records from the analysis.
\begin{Schunk}
\begin{Sinput}
 msmr <- glm((lex.Xst == "Dead") ~ sex - 1 + offset(log(E)),
              family = poisson,
              data = subset(SLr,E>0))
 ci.exp(msmr)
\end{Sinput}
\end{Schunk}
  Recogninze the numbers?
\item The same model can be fitted a bit simpler by the
\code{poisreg} family, try:
\begin{Schunk}
\begin{Sinput}
 msmr <- glm(cbind(lex.Xst == "Dead", E) ~ sex - 1,
             family = poisreg,
               data = subset(SLr, E > 0))
 ci.exp(msmr)
\end{Sinput}
\end{Schunk}
  We can assess the ratios of SMRs between men and women by using the
  \texttt{ctr.mat} argument which should be a matrix:
\begin{Schunk}
\begin{Sinput}
 (CM <- rbind(M = c(1,0),
              W = c(0,1),
          'M/F' = c(1,-1)))
 round(ci.exp(msmr, ctr.mat = CM), 2)
\end{Sinput}
\end{Schunk}
What do you conclude?
\end{enumerate}
\subsection*{Age-specific mortality}
\begin{enumerate}[resume]
\item Now estimate age-specific mortality curves for men and
  women separately, using splines as implemented in \texttt{gam}.
  We use \texttt{k = 20} to be sure to catch any irregularities by age.
\begin{Schunk}
\begin{Sinput}
 r.m <- gam(cbind(lex.Xst == "Dead", lex.dur) ~ s(A, k = 20),
             family = poisreg,
             data = subset(SL, sex == "M"))
 gam.check(r.m)
\end{Sinput}
\end{Schunk}
Make sure you understand all the components on this modeling statement.
Fit the same model for women.
There is a convenient wrapper for this, exploiting the \code{Lexis}
structure of data, but which does not have an update
\begin{Schunk}
\begin{Sinput}
 r.m <- gam.Lexis(subset(SL, sex == "M"), ~ s(A, k = 20))
 r.f <- gam.Lexis(subset(SL, sex == "F"), ~ s(A, k = 20))
 gam.check(r.m)
 gam.check(r.f)
\end{Sinput}
\end{Schunk}
\item Now, extract the estimated rates by using the wrapper function
  \texttt{ci.pred} that computes predicted rates and confidence
  limits for these.
  However, when using the \code{glm.Lexis} or \code{gam.Lexis} we
  aviod this; they rely on the \code{poisreg} family that will return
  the rates in the (inverse) units in which the person-years were
  given; that is the units of \code{lex.dur}.
\begin{Schunk}
\begin{Sinput}
 nd <-  data.frame(A = seq(20,90,0.5))
 p.m <- ci.pred(r.m, newdata = nd)
 p.f <- ci.pred(r.f, newdata = nd)
 str(p.m)
\end{Sinput}
\end{Schunk}
\item Plot the predicted rates for men and women together - using for
  example \texttt{matplot} or \texttt{matshade}.
\begin{Schunk}
\begin{Sinput}
 p.f <- ci.pred(r.f, newdata = nd)
 matplot(nd$A, cbind(p.m,p.f) * 1000,
         type = "l", col = rep(c("blue","red"),each = 3), lwd = c(3,1,1), lty = 1,
         log = "y", xlab = "Age", ylab = "Mortality of DM ptt per 1000 PY")
\end{Sinput}
\end{Schunk}
\end{enumerate}
\subsection*{Further time scales: period and duration}
\begin{enumerate}[resume]
\item We now want to model the mortality rates among diabetes patients
  also including current date and duration of diabetes, using penalized
  splines.  Use the argument \texttt{bs = "cr"} to \texttt{s()} to get
  cubic splines indstead of thin plate (\texttt{"tp"}) splines which is
  the default. Check if you have a reasonable fit using \code{gam.check}.
\begin{Schunk}
\begin{Sinput}
 Mtp <- gam((lex.Xst == "Dead") ~ s(  A, bs = "cr", k = 10) +
                                  s(  P, bs = "cr", k = 10) +
                                  s(dur, bs = "cr", k = 10),
             offset = log(lex.dur),
             family = poisson,
               data = subset(SL, sex == "M"))
\end{Sinput}
\end{Schunk}
An easier specification of the model exploits the \code{Lexis} class
of the dataset, try:
\begin{Schunk}
\begin{Sinput}
 Mcr <- gam.Lexis(subset(SL, sex == "M"),
                  ~ s(  A, bs = "cr", k = 10) +
                    s(  P, bs = "cr", k = 10) +
                    s(dur, bs = "cr", k = 10))
 summary(Mcr)
 gam.check(Mcr)
\end{Sinput}
\end{Schunk}
Fit the same model for women as well. Are the models reasonably fitting?
\item Plot the estimated effects, using the default plot method for
  \texttt{gam} objects. Remember that there are three effects
  estimated, so it is useful set up a multi-panel display, and for
  the sake of comparability to set ylim to the same for men and women:
\begin{Schunk}
\begin{Sinput}
 par(mfrow = c(2,3))
 plot(Mcr, ylim = c(-3,3))
 plot(Fcr, ylim = c(-3,3))
\end{Sinput}
\end{Schunk}
What is the absolute scale for these effects?
\item Compare the fit of the naive model with just age and the
  three-factor models, using \texttt{anova}, e.g.:
\begin{Schunk}
\begin{Sinput}
 anova(Mcr, r.m, test = "Chisq")
\end{Sinput}
\end{Schunk}
What do you conclude?
\item The model we fitted has three time-scales: current age, current
  date and current duration of diabetes, so the effects that we report
  are not immediately interpretable, as they are (as in any kind of
  multiple regressions) to be interpreted as ``all else equal'' which
  they are not, as the three time scales advance simultaneously at the
  same pace.
  The reporting would therefore more naturally be on the
  mortality scale as a function of age, but showing the mortality
  for persons diagnosed in different ages, using separate displays
  for separate years of diagnosis.
  This is most easily done using the \texttt{ci.pred} function with
  the \texttt{newdata = } argument. So a person diagnosed in age 50 in
  1995 will have a mortality measured in cases per 1000 PY as:
\begin{Schunk}
\begin{Sinput}
 pts <- seq(0, 12, 1/4)
 nd <- data.frame(A =   50 + pts,
                  P = 1995 + pts,
                dur =        pts)
 m.pr <- ci.pred(Mcr, newdata = nd)
\end{Sinput}
\end{Schunk}
  Note that because we used \code{gam.Lexis} which uses
  the\code{poisreg} family we need not specify \code{lex.dur} as a
  variable in the prediction data frame \code{nd}. Predictions will
  be rates in the same units as \code{lex.dur}.
  Now take a look at the result from the \texttt{ci.pred} statement and
  construct prediction of mortality for men and women diagnosed in a
  range of ages, say 50, 60, 70, and plot these together in the same
  graph:
\begin{Schunk}
\begin{Sinput}
 cbind(nd, ci.pred(Mcr, newdata = nd))
\end{Sinput}
\end{Schunk}
\item From figure it seems that the duration effect is
over-modeled, so refit constraining the d.f. to 5:
\begin{Schunk}
\begin{Sinput}
 Mcr <- gam.Lexis(subset(SL, sex == "M"),
                  ~ s(  A, bs = "cr", k = 10) +
                    s(  P, bs = "cr", k = 10) +
                    s(dur, bs = "cr", k = 5))
 summary(Mcr)
 Fcr <- gam.Lexis(subset(SL, sex == "F"),
                  ~ s(  A, bs = "cr", k = 10) +
                    s(  P, bs = "cr", k = 10) +
                    s(dur, bs = "cr", k = 5))
\end{Sinput}
\end{Schunk}
How does \texttt{gam.check()} look for these models?
Plot the estimated rates from the revised models.
What do you conclude from the plots?
\end{enumerate}
\subsection{SMR modeling}
\begin{enumerate}[resume]
\item Now model the SMR using age and date of diagnosis and diabetes
  duration as explanatory variables, including the expected-number
  instead of the person-years, using separate models for
  men and women.
  You cannot use \texttt{gam.Lexis} from the code you used for
  fitting models for the rates, you need to use \texttt{gam} with
  the \texttt{poisreg} family. And remember to exclude those units
  where no deaths in the population occur (that is where the rate
  is 0).
  Plot the estimated smooth effects for both men and women using
  e.g. \texttt{plot.gam}. What do you see?
\item Plot the predicted SMRs from the models for men and women
  diagnosed in ages 50, 60 and 70 as you dif for the rates. What do
  you see?
\item Try to simplify the model to one with a simple sex effect,
  separate linear effects of age and date of follow-up for each
  sex, and a smooth effect of duration common for both sexes,
  giving an estimate of the change in SMR by age and calendar
  time. How much does SMR change by each year of age? And by each
  calendar year?
\item Use your previous code to plot the predicted mortality from this
  model too. Are the predicted SMR curves credible?
\end{enumerate}
