---
title: "SPE-R software requirement"
output: 
  html_document: 
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Software required

## R version

The course material have been compiled and tested with
`r R.version$version.string`.
We advise to install this version of `R`. For installation instructions please refer to [CRAN wbsite](https://cran.r-project.org/).

## Code source editing software

Any code editing software can be used during the course.
[RStudio](https://www.rstudio.com/) is an adapted solution. 

## R packages

```{r pkgs list, include=FALSE}
## install checkpoint package if not available
if(!require(checkpoint)){ 
  install.packages('checkpoint') 
  library(checkpoint)
}

if(!require(purrr)){ 
  install.packages('rlang') 
  library(rlang)
}


required.pkg <- checkpoint:::scanForPackages(project = '../pracs/')

required.pkg
paste0(sort(required.pkg$pkgs), collapse = ', ')
command.str <- 
  paste0(
    '```r \n', 
    'install.package(c("', 
    paste0(sort(required.pkg$pkgs), collapse = '", "'), 
    '")) \n',
    '```\n')

```

During the practical session, the following list of `R`the packages will be required:

`r paste0('- ', paste(required.pkg$pkgs, collapse = "\n- "))`

which can be installed with the following command:

`r command.str`

---
[back to mainpage](https://spe-r.github.io/)
