---
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
knitr::opts_chunk$set(results = "hide", fig.show="hide", messages=FALSE, keep.source = TRUE, include = TRUE, eps = FALSE, prefix.string = "./graph/causInf2")
```

<!-- ```{r include=FALSE, warning=FALSE, message=FALSE} -->

<!-- base.pkg.list <- c("stats", "graphics", "grDevices", "datasets", "utils", "methods", "base") -->

<!-- loaded.pkg.list <- setdiff((.packages()), base.pkg.list) -->

<!-- for(pkg_ in loaded.pkg.list) { -->

<!--   detach(paste0("package:", pkg_), unload = TRUE, character.only = TRUE) -->

<!-- } -->

<!-- ``` -->

# Causal inference 2: Model-based estimation of causal parameters

Sources of inspiration when creating this practical: 
[Luque Fernandez, M.A. et al. (2018)](https://doi.org/10.1002/sim.7628)
*Stat Med*
2018;37(16):2530-2546 and
[Smith et al. (2022)](https://doi.org/10.1002/sim.9234) *Stat Med*
2022;41(2):407-432.

## Introduction

In this exercise we learn to apply some R tools for estimating
interesting causal parameters in the context of an observational study.
The statistical methods to be introduced are (a) regression
standardization or g-formula, (b) inverse probability weighting (IPW)
with propensity scores and (c) double robust (DR) estimation combining
approaches (a) and (b).

We illustrate these methods using simulated data. The exposure variable
$X$ is binary (1/0) and the outcome variable $Y$ is binary, too.\
The interesting estimands are the causal risk difference (RD) and the
causal risk ratio (RR).

As a background story, we imagine a population of cancer patients, in
whom the exposure, and four covariates $Z = (Z_1, Z_2, Z_3, Z_4)$ plus
the assumed marginal distributions of the latter are described in the
table below:

| Variable | Description |
|:-----------------|:-----------------------------------------------------|
| $X$ | treatment; 1: radiotherapy only, 0: radiotherapy + chemotherapy |
| $Y$ | death during one year after diagnosis of cancer |
| $Z_1$ | sex; 0: man, 1: woman; $Z_1 \sim \text{Bern}(0.5)$ |
| $Z_2$ | age group 0; `young`, 1: `old`; $Z_2 \sim \text{Bern}(0.65)$ |
| $Z_3$ | stage of cancer; 4 classes; $Z_3 \sim \text{DiscUnif}(1, \dots, 4)$ |
| $Z_4$ | comorbidity score; 5 classes; $Z_3 \sim \text{DiscUnif}(1, \dots, 5)$ |

For simplicity, covariates $Z_3$ and $Z_4$ are treated as continuous
variables with linear effects in the models.

1.  First, load the packages needed in this exercise

```{r packages}
library(dagitty)
library(Epi)
library(stdReg2)
library(PSweight)
```

2.  The assumed causal diagram is drawn using `dagitty` and is shown
    below.

```{r dagitty, echo=FALSE, message=FALSE, warning=FALSE, results=TRUE, fig.show=TRUE}
diagram <-
  dagitty("dag {
  Z2 -> Z3 -> Y
  Z2 -> Z4 -> Y
  Z2 -> Y
  Z2 -> Z3 -> X
  Z2 -> Z4 -> X
  Z2 -> X
  Z1 -> Z3 -> Y
  Z1 -> Z4 -> Y
  Z1 -> Y
  Z1 -> Z3 -> X
  Z1 -> Z4 -> X
}")

coordinates(diagram) <-
  list(
    x = c(X = 1, Y = 5, Z1 = 5, Z2 = 1, Z3 = 4, Z4 = 2),
    y = c(X = 2, Y = 2, Z1 = 0, Z2 = 0, Z3 = 1, Z4 = 1)
  )

plot(diagram)
```

3.  For a more generic notation, the risk of death, i.e. the probability
    of $Y=1$, will be expressed as the expected value of the binary
    outcome $Y$, for instance
    $$ E(Y^{X=x}) = P(Y^{X=x}=1) \quad \text{and} \quad 
    E(Y|X=x, Z=z) = P(Y=1|X=x, Z=z), $$ where $Z$ is the vector of
    relevant covariates.

The same principle is applied in expressing the conditional probability
of being exposed, i.e. that of $X=1$, given a set of covariate values
$Z=z$: $$ E(X|Z=z) = P(X=1|Z=z). $$ The fitted and predicted
probabilities of $Y=1$ are denoted as $\widehat{Y}$ and $\widetilde{Y}$,
respectively, with pertinent subscripts and/or superscripts.

Both $X$ and $Y$ are modelled by logistic regression. The expit-function
or inverse of the logit function is defined:
$\text{expit}(u) = 1/(1 + e^{-u})$, $u\in R$. This is equal to the
cumulative distribution function of the standard logistic distribution,
the values of which are returned in R by `plogis(u)`. The R function
that returns values of the logit-function is `qlogis()`.

The **true exposure model** for the dependence of exposure $X$ on the
covariates $Z$ is specified as $$ E(X|Z_1 = z_1, \dots, Z_4 = z_4) =
      \text{expit}(-5 + 0.05z_2 + 0.25z_3 + 0.5z_4 + 0.4z_2z_4) . $$ The
assumed **true outcome model**
$$ E(Y|X=x, Z_1 = z_1, \dots, Z_4 = z_4) =
           \text{expit}(-1 + x - 0.1z_1 + 0.35z_2 + 0.25z_3 +
                 0.20z_4 + 0.15z_2z_4) $$ Note that $X$ does not depend
on $Z_1$, and that in both models there is a product term $Z_2 Z_4$, the
effect of which appears weaker in the outcome model.

## Control of confounding

1.  Based on inspection of the causal diagram, can you provide
    justification for the claim that variables $Z_2, Z_3$ and $Z_4$ form
    a proper subset of the four covariates, which is sufficient to block
    all backdoor paths between $X$ and $Y$ and thus remove confounding?

2.  Look what dagitty offers for sufficient sets:

```{r suff sets, echo=FALSE, message=FALSE, warning=FALSE}
adjustmentSets(diagram, exposure="X", outcome="Y")
```

3.  Even though we have a sufficient set as claimed above, why it still
    might be useful to include covariate $Z_1$, too, when modelling the
    outcome?

## Generation of target population and true models

1.  Define two R-functions, which compute the expected values for the
    exposure and those for the outcome based on the assumed true
    exposure model and the true outcome model, respectively.

```{r true models}
EX <- function(z2, z3, z4) {
  plogis(-5 + 0.05 * z2 + 0.25 * z3 + 0.5 * z4 + 0.4 * z2 * z4)
}
EY <- function(x, z1, z2, z3, z4) {
  plogis(-1 + x - 0.1 * z1 + 0.35 * z2 + 0.25 * z3 +
    0.20 * z4 + 0.15 * z2 * z4)
}
```

2.  Define a function for generation of data by simulating random values
    from pertinent probability distributions based on the given
    assumptions.

```{r data generation function}
genData <- function(N) {
  z1 <- rbinom(N, size = 1, prob = 0.5) # Bern(0.5)
  z2 <- rbinom(N, size = 1, prob = 0.65) # Bern(0.65)
  z3 <- trunc(runif(N, min = 1, max = 5), digits = 0) # DiscUnif(1,4)
  z4 <- trunc(runif(N, min = 1, max = 6), digits = 0) # DiscUnif(1,5)
  x <- rbinom(N, size = 1, prob = EX(z2, z3, z4))
  y <- rbinom(N, size = 1, prob = EY(x, z1, z2, z3, z4))
  data.frame(z1, z2, z3, z4, x, y)
}
```

3.  Generate a data frame `pop` for a big target population of 1000000
    subjects

```{r popdata generation}
N <- 1000000
set.seed(7777)
pop <- genData(N)
```

## Factual and counterfactual risks - associational and causal contrasts

1.  Compute the factual risks of death for the two exposure groups
    $$ E(Y|X=x) = P(Y=1|X=x) = \frac{P(Y=1\ \&\ X=x)}{P(X=x)},
    \quad x=0,1, $$ 
    in the whole target population, as well as their
    associational contrasts: risk difference, risk ratio, and odds
    ratio. Before that let us define a useful function `Contr()`:

```{r Contr function and associational contrasts}
Contr <- function(mu1, mu0) {
  RD <- mu1 - mu0
  RR <- mu1 / mu0
  OR <- (mu1 / (1 - mu1)) / (mu0 / (1 - mu0))
  return(c(Risk1=mu1, Risk0=mu0, RD = RD, RR = RR, OR = OR))
}
Ey1fact <- with(pop, sum(y == 1 & x == 1) / sum(x == 1))
Ey0fact <- with(pop, sum(y == 1 & x == 0) / sum(x == 0))
round(Contr(Ey1fact, Ey0fact), 4)
```

The one-year mortality is quite high in both groups. -- How much bigger
is the risk of death of those factually exposed to radiotherapy only as
compared with those receiving chemotherapy, too?

*Comment*: With such a high baseline probability of the outcome the risk
ratio cannot become greater than 1/0.628 = 1.59. For relative
comparisons one could also consider using **survival ratio** (SR) For
instance, here the associational survival ratio is
$$ \text{SR} = (1-0.8955)/(1-0.6280) = 0.1045/0.372 = 0.2809 = 1/3.56 . $$
See also [Huitfeldt et al. (2022)](https://doi.org/10.48550/arXiv.2106.06316) 
"Shall we count the living or the dead?'' 
for discussion concerning the choice between risk ratio and survival
ratio for a relative effect measure.

2.  Compute now the true **counterfactual** or **potential risks** of
    death $$ E(Y_i^{X_i=x}) = P(Y_i^{X_i=x}=1) = \pi_i^{X_i=x} $$ for
    each individual under the alternative treatments or exposure values
    $x=0,1$ with given covariate values. After that derive the average
    or overall counterfactual risks $E(Y^{X=1}) = \pi^1$ and
    $E(Y^{X=0}) = \pi^0$ in the population, and finally the true
    **marginal causal contrasts** describing the effect of $X$:
    $$\begin{aligned}
     \text{RD} & = E(Y^{X=1})-E(Y^{X=0}), \qquad  
     \text{RR} = E(Y^{X=1})/E(Y^{X=0}), \\
     \text{OR} & = 
     \frac{E(Y^{X=1})/[1 -  E(Y^{X=1})]}{E(Y^{X=0})/[1 -  E(Y^{X=0})] }
    \end{aligned} $$

```{r true contrasts}
pop <- transform(pop,
  EY1.ind = EY(x = 1, z1, z2, z3, z4),
  EY0.ind = EY(x = 0, z1, z2, z3, z4)
)
EY1pot <- mean(pop$EY1.ind)
EY0pot <- mean(pop$EY0.ind)
round(Contr(EY1pot, EY0pot), 4)
```
See [Colnet et al. (2023)](https://doi.org/10.48550/arXiv.2303.16008) 
for a thorough 
comparison of the different causal contrasts. Among these,
RD appears to have more favourable mathematical properties
than RR and OR, and in several respects OR is a 
particularly problematic measure of causal effect.

3.  Compare the associational contrasts (e.g. RD=0.2674) computed in
    item 1 with the causal contrasts here. What do you conclude about
    the amount of confounding bias in the associational contrasts?

4.  To illustrate what happens in a realistic setting involving some
    random variability we shall draw a random sample of 5000 patients
    using systematic sampling. We also compute the estimates of the
    counterfactual quantities from this sample based on the true models.

```{r sample}
n <- 5000
samp <- pop[seq(1,N, by=N/n), ]
str(samp)
EY1pot.samp <- mean(samp$EY1.ind)
EY0pot.samp <- mean(samp$EY0.ind)
round(Contr(EY1pot.samp, EY0pot.samp), 4)
```

The sample estimates of the counterfactual risks and their contrasts
based on the true models seem to be not too far from the true values of
those estimands.

5.  Some perspective to the random variability in the estimation of
    parameters of interest is provided by the numbers of cases and
    non-casses in the two exposure groups of the sample

```{r cases by exposure}
stat.table(index=list("Outcome"=factor(y), "Exposure"=factor(x)),
           contents=list(count(), percent(y)),
           margins=TRUE, data=samp  )
```

The group sizes are very different; only 16 % of the patients were
exposed to radiotherapy only.

The Statistical precision in any comparison between the exposure groups
based on the sample data is very much dominated by the smallest count in
this fourfold table, i.e. the number of non-cases among the exposed,
which is not very big.

## Outcome modelling and estimation of causal contrasts by g-formula

As the first approach for estimating causal contrasts of interest we
apply the method of **regression standardization** or **g-formula**. It
is based on a hopefully realistic enough model for $E(Y|X=x, Z=z)$, i.e.
how the risk of the outcome is expected to depend on the exposure
variable $X$ and on a sufficient set $Z$ of confounders. The potential
or counterfactual risks $E(Y^{X=x}), x=0,1$, are marginal expectations
of the above quantities, standardized over the joint distribution of the
confounders $Z$ in the target population.
$$ E(Y^{X=x}) = E_Z[E(Y|X=x,Z)]
       = \int E(Y|X=x, Z=z)dF_Z(z), \quad x=0,1. $$

1.  Assume now a *slightly misspecified* model `mY` for the outcome,
    which contains only the main effect terms of the explanatory
    variables:
    $$ \pi_i = E(Y_i|X_i=x_i, Z_{i1}=z_{i1}, \dots, Z_{i4}=z_{i4}) =
      \text{expit}\left(\beta_0 + \delta x_i +
      \sum_{j=1}^4 \beta_j z_{ij} \right) . $$ Fit this model on the
    data of the whole target population using function `glm()`
    <!-- % in order to have an accurate -->
    <!-- % estimate of the possible bias due 
    to misspecification of the outcome --> <!-- % model -->

```{r outcome model}
mY <- glm(y ~ x + z1 + z2 + z3 + z4, family = binomial, data = pop)
round(ci.lin(mY, Exp = TRUE)[, c(1, 5:7)], 4)
```

The point estimates for the intercept, $X, Z_1$ and $Z_3$ seem to be
reasonably close to the given parameter values in the true model, but
those of $Z_2$ and $Z_4$ are different. This is because the fitted model
ignored the product term $Z_2 Z_4$.

2.  For each subject $i$, compute the **fitted individual risk**
    $\widehat{Y_i}$ (which actually will be needed not until the end of
    this practical!) as well as the **predicted potential
    (counterfactual) risks** $\widetilde{Y_i}^{X_i=x}$ for both exposure
    levels $x=0,1$ separately, keeping the individual values of the
    $Z$-variables as they are observed.

```{r fitted risks and predicted potential risks}
pop$yh <- predict(mY, type = "response") #  fitted values
pop$yp1 <- predict(mY, newdata = data.frame(
  x = rep(1, N), # x=1
  pop[, c("z1", "z2", "z3", "z4")]
), type = "response")
pop$yp0 <- predict(mY, newdata = data.frame(
  x = rep(0, N), # x=0
  pop[, c("z1", "z2", "z3", "z4")]
), type = "response")
```

3.  Applying the method of standardization or g-formula we now compute
    the point estimates $$ \widehat{E}_g(Y^{X=x}) =
     \frac{1}{n} \sum_{i=1}^n \widetilde{Y}_i^{X_i=x}, \quad x=0,1. $$
    of the two counterfactual risks $E(Y^{X=1}) = \pi^1$ and
    $E(Y^{X=0})=\pi^0$ as well as\
    those of the marginal causal contrasts.

```{r causal contrasts}
EY1pot.g <- mean(pop$yp1)
EY0pot.g <- mean(pop$yp0)
round(Contr(EY1pot.g, EY0pot.g), 4)
```

*Comment*: The joint distribution of the confounders $Z$ in the target
population is here represented by the actual distribution of the data in
that population. Thus, marginal expectations $E_Z[E(X=x, Z)]$ are
obtained by simple computation of the arithmetic means of the
individually predicted values $\widetilde{Y_i}^{X_i=x}$ of the outcome
for the two exposure levels.

Compare the estimated contrasts with those obtained from applying the
true model (RD = 0.1743) in item 4.4 above. How big is the bias due to
misspecification of the outcome model?

Compare in particular the estimate of the marginal OR here with the
conditional OR obtained in item 5.1 from the pertinent coefficient in
the logistic model. Which one is closer to 1?

4.  Perform the same calculations using the tools in package `stdReg2`
    but now on the sample data. (see [Sjölander
    2016](https://doi.org/10.1007/s10654-016-0157-3) and [Sachs et al
    2025](https://doi.org/10.1016/j.annepidem.2025.04.002)). Using
    function `standardize_glm()` the same logistic model is fitted, and
    the point estimates together with the confidence intervals for the
    causal risk difference and causal risk ratio are computed. The
    results are saved into a list, for which we give the name `mY.std`.

The main results are shown in four small windows\
after applying the basic print method on `mY.std`

```{r stdReg2}
mY.std <- standardize_glm(
 formula = y ~ x + z1 + z2 + z3 + z4,
 family = "binomial",
 data = samp,
 values = list(x = 0:1),
 contrasts = c("difference", "ratio"),
 reference = 0
)
mY.std
```

The point estimates seem to be well within a resonable error margin when
compared to those, which you got by explicit computations for the whole
population data (e.g. RD=0.1828) in the previous item. (You could also
run `standardize_glm()` with `data=pop` to convince yourself that
`stdReg2' indeed provides the same point estimates  as obtained above for the whole population from model`mY\`).

In `stdReg2`, the standard errors are obtained by the multivariate delta
method built upon M-estimation ([Stefanski & Boos
2002](https://doi.org/10.1198/000313002753631330)) and robust sandwich
estimator of the pertinent covariance matrix. Approximate confidence
intervals are derived from these in the usual way.

## Inverse probability weighting (IPW) by propensity scores

Our next method is based on weighting each individual observation by the
inverse of the probability of belonging to that particular exposure
group, which was realized, this probability being predicted by the
determinants of exposure.

1.  First, we examine the validity of this approach by fitting the true
    model on the whole population data.\
    To repeat, the model is $$ 
     p_i = E(X_i| Z_{1i} = z_{1i}, \dots, Z_{4i} = z_{4i})
     = \text{expit}(\gamma_0 + \gamma_2 z_{2i} +
      \gamma_3 z_{i3} + \gamma_4 z_{4i} + \gamma_{24} z_{2i}z_{4i}), \quad i=1, \dots N 
     $$

```{r exposure model}
mX <- glm(x ~ z2 + z3 + z4 + z2:z4,
  family = binomial(link = logit), data = pop)
round(ci.lin(mX, Exp = TRUE)[, c(1, 5)], 4)
```

The estimated coefficients and ORs are quite close to the true values as
given in the introduction section.

2.  From the fitted model we then extract the **propensity scores**,
    i.e. fitted probabilities of belonging to exposure group 1:
    $\text{PS}_i = \widehat{p_i}$. We also make a simple comparison of
    the distribution of the scores between the two groups by common
    summary measures.

```{r propScore, fig=FALSE}
pop$PS <- predict(mX, type = "response")
tapply(pop$PS, pop$x, summary)
```

How different are the distributions? Can one judge from these summaries,
that they are sufficiently overlapping?

Informative graphical tools for examining overlap and covariate balance
are contained in package `PSweight`,\
and they are introduced in the next section.

3.  We then compute the **inverse probability weights** (IPW), also
    known as **inverse probability treatment weights** (IPTW):
    $$\begin{aligned}
     W_i & = \frac{1}{\text{PS}_i}, \quad \text{when }\ X_i=1, \\
     W_i & = \frac{1}{1-\text{PS}_i}, \quad \text{when }\ X_i=0 . 
    \end{aligned} $$ Look at the sum as well as the distribution summary
    of the weights in the exposure groups. In an ideal situation with a
    well-specified exposure model the sum of weights should be close to
    to the population size in both groups.

```{r ipw weights}
pop$w <- ifelse(pop$x == 1, 
                     1 / pop$PS, 1 / (1 - pop$PS))
with(pop, tapply(w, x, sum))
```

4.  Compute now the weighted estimates of the potential or
    counterfactual risks for both exposure categories
    $$ \widehat{E}_w(Y^{X = x}) =
    \frac{ \sum_{i=1}^n {\mathbf 1}_{ \{X_i=x\} } W_i Y_i }
       {\sum_{i=1}^n {\mathbf 1}_{ \{X_i=x\} }W_i} =
     \frac{ \sum_{X_i = x} W_i Y_i }{\sum_{X_i=x} W_i}, \quad x = 0,1, $$
    and their causal contrasts, for instance
    $$ \widehat{\text{RD}}_{w} = \widehat{E}_w(Y^{X = 1}) -
                  \widehat{E}_w(Y^{X = 0})
      =  \frac{ \sum_{i=1}^n X_i W_i Y_i }{\sum_{i=1}^n X_i W_i} -
      \frac{ \sum_{i=1}^n (1-X_i) W_i Y_i }{\sum_{i=1}^n (1-X_i) W_i}$$

```{r ipw-estimate}
EY1pot.w <- sum(pop$x * pop$w * pop$y) / sum(pop$x * pop$w)
EY0pot.w <- sum((1 - pop$x) * pop$w * pop$y) / sum((1 - pop$x) * pop$w)
round(Contr(EY1pot.w, EY0pot.w), 4)
```

These results seem to be practically equal to the true values (e.g. true
RD=0.1743).

*Comment*: If you apply the methods introduced in the next section but
use the whole population data, i.e. putting `data=pop`, you would see
that the estimates of the causal contrasts would be the same as obtained
here.

## IPW estimation using R package `PSweight`

We shall now try IPW-estimation with a misspecified model, such that we
include all $Z$-variables but none of their pairwise interactions. We do
this on the sample data only. In computations we utilize the R package
`PSweight` (see [PSweight
vignette](https://cran.rstudio.com/web//packages/PSweight/vignettes/Software_vig.pdf)).

1.  The exposure model is fitted using `glm()`, and the weights are
    obtained using function `SumStat()` in package `PSweight`:

```{r PSweight, fig=FALSE}
mX2 <- glm(x ~ z1 + z2 + z3 + z4, family = binomial, data = samp)
round(ci.lin(mX2, Exp = TRUE)[, c(1:2, 5:7)], 3)
psw2 <- SumStat(
  ps.formula = mX2$formula, data = samp,
  weight = c("IPW", "treated", "overlap")
)
```

Note that apart from weights `IPW` for the whole population, other kinds
of weights can also be computed. These are relevant when estimating
other types of causal contrasts, like "average treatment effect among
the treated" (ATT, see below) and "average treatment effect in the
overlap (or equipoise) population" (ATO).

2.  `PSweight` includes some useful tools to examine the properties of
    the distribution of the propensity scores. First we draw density
    plots showing, how well the distributions of propensity scores (and
    1 - PS, too) are overlapping between the exposure groups -- **NB.**
    You may get a warning about usage of a\
    `ggplot2` aesthetic in the density plot implemented in `PSweight`,
    this feature being now outdated in the newest version of `ggplot2`.
    Nonetheless, the density plot will after all be drawn, if you go
    down on the Console window and "Press [enter] to continue"!

```{r distributions of the scores, fig=FALSE}
plot(psw2, type = "density")
```

Are the distributions of the propensity scores very different?

3.  We then create a diagnostic plot to examine the covariate balance
    between the groups:

```{r check balance, fig=FALSE}
plot(psw2, type = "balance")
```

In this plot, it would be desirable that the horisontal values of the
standardized mean differences for a given type of weights are less than
0.1. -- There seems to be some issue with the balance in the estimation
of our target parameters based on this exposure model.

4.  Estimation and reporting of the causal contrasts: Weights must be
    `IPW` when estimating the causal effect generalized to the whole
    target population. For relative contrasts, the summary method
    provides the results on the log-scale; therefore
    $\exp$-transformation.

```{r ipw-estimation}
ipw2est <- PSweight(ps.formula = mX2, yname = "y", data = samp, 
                    weight = "IPW")
ipw2est
summary(ipw2est, type="DIF")  # risk difference
(logRR.ipw2 <- summary(ipw2est, type = "RR"))  # log(risk ratio)
round(exp(logRR.ipw2$estimates[c(1, 4, 5)]), 3) # risk ratio
round(exp(summary(ipw2est, type = "OR")$estimates[c(1, 4, 5)]), 3) # OR
```

Comparing these with the previous IPW estimates (e.g. RD=0.1737) as well
as the true values (RD=0.1743), there perhaps can be seen indication of
some bias due to misspecification. Remember, though, that these
estimates are from the sample data with relatively wide error margins.

The standard errors provided by `PSweight` are by default based on the
empirical sandwich covariance matrix and application of delta method as
appropriate. Bootstrapping is also possible but is computationally very
intensive and is recommended to be used only in relatively small
samples.

## Effect of exposure among those exposed

If we are interested in the causal contrasts describing the **effect of
exposure among those exposed** (like ATE), the relevant factual and
counterfactual risks in that subset are conditional on being exposed:

$$\begin{aligned}
 \pi^1_1 & = E(Y^{X=1}|X=1) = E(Y|X=1) = \pi_1, \\
 \pi^0_1 & = E(Y^{X=0}|X=1) = \sum_{X_i=1} E(Y|X=0, Z=z)P(Z=z|X=1)
\end{aligned}$$

We are thus making and "observed vs. expected" comparison, in which the
$z$-specific risks in the unexposed are weighted by the distribution of
$Z$ in the exposed subset of the target population.

1.  We shall first estimate these $z$-specific "observed" or factual and
    "expected" or counterfactual risks and their contrasts conditional
    on being exposed from the slightly misspecified outcome model `mY`
    fitted on the whole population data:

```{r g-formula-att}
EY1att.g <- mean(subset(pop, x == 1)$yp1)
EY0att.g <- mean(subset(pop, x == 1)$yp0)
round(Contr(EY1att.g, EY0att.g), 4)
```

Compare the results here with those (e.g. RD=0.1828) meant to describe
the overall causal contrasts among the whole target population from the
same model `mY`. What do you observe?

2.  What is your guess about the causal effect of exposure among the
    unexposed; is it bigger or smaller than it is among the exposed or
    among the whole population?

3.  Incidentally, the true causal contrasts among the exposed based on
    the true model are similarly obtained from the quantities in item
    4.2 above:

```{r true among exposed}
EY1att <- mean(subset(pop, x == 1)$EY1.ind)
EY0att <- mean(subset(pop, x == 1)$EY0.ind)
round(Contr(EY1att, EY0att), 4)
```

Compare the estimates in the previous item (e.g. RD$_1$ = 0.1389) with
the true values obtained here. The misspecification of the outcome model
implies some bias in these results, too.

4.  When wishing to estimate the effect of exposure among the exposed
    using the IPW approach, then the weights are $W_i = 1$ for the
    exposed and $W_i = \text{PS}_i/(1-\text{PS}_i)$ for the unexposed.
    We shall again use `PSweight` with sample data but with another
    choice of weights:

```{r ps-estimation-att}
psatt <- PSweight(ps.formula = mX2, yname = "y", 
                  data = samp, weight = "treated")
psatt
round(summary(psatt)$estimates, 4)
round(exp(summary(psatt, type = "RR")$estimates), 3)
```

Compare the results here with those obtained by g-formula in item 8.1
and with the true contrasts above.

## Double robust (DR) estimation

We may try to reduce the bias in the estimation of the target
parameters, which would be caused by possible specification error either
in the outcome model or in the exposure model using some **double
robust** (DR) approach. These methods combine the principle of
regression standardization with that of IPW/propensity scores.

A classical DR method is *augmented IPW estimation* (AIPW), in which the
estimator can be expressed in two alternative ways: either an
IPW-corrected g-formula estimator, or an IPW-estimator corrected by
standardization.

$$\begin{aligned}
 \widehat{E}_a(Y^{X=x}) 
       & = \widehat{E}_g(Y^{X=x}) +
\frac{1}{n} \sum_{i=1}^n  {\mathbf 1}_{\{X_i=x\}} W_i 
                   ( Y_i - \widetilde{Y}_i^{X_i=x} )   \\
       & = \widehat{E}_w(Y^{X=x}) +
\frac{1}{n} \sum_{i=1}^n ( 1 - {\mathbf 1}_{\{X_i=x\}} W_i ) 
                      \widetilde{Y}_i^{X_i=x}.
\end{aligned}$$

1.  We apply the AIPW method first in the whole population by direct
    combination of the results that we obtained from the slightly
    misspecified outcome model `mY` with those from the well-specified
    exposure model `mX`.

```{r aipw}
EY1pot.a <- EY1pot.g + mean( 1*(pop$x==1) * pop$w * (pop$y - pop$yp1) )
EY0pot.a <- EY0pot.g + mean( 1*(pop$x==0) * pop$w * (pop$y - pop$yp0) )
round(Contr(EY1pot.a, EY0pot.a), 4)
```

Compare these results with those obtained by g-formula (e.g. RD=0.1828)
and by the IPW method (RD=0.1737) as well as with the true values (RD =
0.1743) of the parameters. -- Was correction of the g-estimates by IPW
successful?

2.  AIPW-estimates and confidence intervals for the causal contrasts of
    interest can be obtained, for instance, using `PSweight`. We will
    assign the outcome model `mY` into the argument `out.formula`. We
    again combine this slightly misspecified outcome model with the
    correct exposure model `mX` and perform the estimation on the sample
    data.

```{r AIPW by PSweight}
aipw.psw <- PSweight(ps.formula = mX, out.formula = mY, yname = "y",
                     data = samp, weight = "IPW")
aipw.psw
round(summary(aipw.psw)$estimates, 4)
round(exp(summary(aipw.psw, type = "RR")$estimates), 3)
```

The estimates here are quite close to the true values, but the error
margins are still relatively wide.

3.  Alternatively, `stdReg2` can also be utilized for double-robust
    estimation of causal contrasts. It leans, though, on a somewhat
    different approach. The method developed for the types of outcomes
    and models covered by generalized linear models (GLM) is known by
    its initialism **IPTW GLM**. It is based on weighted maximum
    likelihood fitting of the outcome model. The weights are inverses of
    the propensity scores obtained from the pertinent exposure model.
    Computation of standard errors and confidence intervals uses the
    influence function approach, which among other things takes into
    account the fact that weighting is based on estimated propensity
    scores. -- See [Gabriel et al
    2024](https://doi.org/10.1002/sim.9969) for details of IPTW GLM.

We shall try this method for the sample data with the script below:

```{r aipw by stdReg2}
dr.std <- standardize_glm_dr(
          formula_outcome = y ~ x + z1 + z2 + z3 + z4, 
          formula_exposure = x ~ z2 + z3 + z4 + z2:z4,
          data = samp,
          family_outcome = "binomial", 
          family_exposure = "binomial",
          values = list(x = 0:1), 
          contrasts = c("difference", "ratio"), reference = 0)
dr.std
```

The point estimates are not much different from those obtained above
when applying AIPW estimation.

## DR estimation with flexible learning algorithms

It may not be easy to specify a conventional generalized linear models
or even a generalized additive model for exposure and outcome, which
would be\
sufficiently realistic, yet not suffering from overfitting. Modern
approaches of **statistical learning**, a.k.a. "machine learning"
provide tools for flexible modelling, which may be used to reduce the
risk of misspecification, if thoughtfully applied.

There are a few R packages in which some general algorithmic approaches
for supervised learning are implemented for estimating causal
parameters. For instance, package `AIPW` (see [Zhong et al.,
2021](https://doi.org/10.1093/aje/kwab207)) utilizes several learning
algorithms for exposure and outcome modelling and then performs AIPW
estimation of the parameters of interest coupled with calculation of
confidence intervals. Package `tmle` (see [Karim and Frank,
2021](https://ehsanx.github.io/TMLEworkshop/)) performs same tasks but
uses the **targeted maximum likelihood estimation** (TMLE) approach in
estimation.

Both `AIPW` and `tmle` lean on the `SuperLearner` package, which uses
multiple learning algorithms (e.g. GLM, GAM, Random Forest, Recursive
Partitioning, Gradient Boosting, etc.) for constructing predictions of
the counterfactual quantities, and then creates an optimal weighted
average of those models, a.k.a. an "ensemble". These algorithms are
computationally highly intensive. Fitting models with only 3 or 4
covariates as in this practical for our whole target cohort of 1,000,000
subjects would take hours on an ordinary laptop. With a sample of 5000
it woul still take several minutes.

## Extra: Computations in the TMLE method

Finally, a few words on, how causal contrasts are estimated using the
TMLE, when we have fitted an outcome model, like `mY`, and an exposure
model, like `mX`. TMLE corrects the estimates obtained from the outcome
model by some elements derived from the exposure model. The correction,
though, is not as intuitive and transparent as it is in AIPW. See
[Schuler and Rose (2017)](https://doi.org/10.1093/aje/kww165) for more
details.

We show the steps of computation using the data covering the whole
target population.

1.  In the first step we utilize the propensity scores previously
    obtained from fitting the correct exposure model and compute from
    them the "**clever covariates**" `H1`and `H0`:

```{r clever covariates}
pop$H1 <- pop$x / pop$PS
pop$H0 <- (1 - pop$x) / (1 - pop$PS)
```

2.  Then, a **working model** is fitted for the outcome, in which the
    clever covariates are explanatory variables, but the model also
    includes as an offset term the previously fitted linear predictor
    from the original outcome model `mY`:
    $\widehat{\eta}_i = \text{logit}(\widehat Y_i)$ based on the
    observed exposure $X_i=x$ for each subject $i$, where the fitted
    values $\widehat Y_i$ were saved into vector `d$yh` already in item
    5.2 above. Moreover, the intercept is removed from the model.

```{r model with clever covariates}
epsmod <- glm(y ~ -1 + H0 + H1 + offset(qlogis(yh)),
  family = binomial(link = logit), data = pop)
eps <- coef(epsmod)
eps
```

3.  The logit-transformed predicted values $\widetilde{Y}_i^{X_i=1}$ and
    $\widetilde{Y}_i^{X_i=0}$ of the potential or counterfactual
    individual risks from the original outcome model `mY` are now
    corrected by the estimated coefficients of the clever covariates,
    and the corrected predictions are returned to the original
    probability scale.

```{r tmle predictions}
pop$ypred0.H <- plogis(qlogis(pop$yp0) + eps[1] / (1 - pop$PS))
pop$ypred1.H <- plogis(qlogis(pop$yp1) + eps[2] / pop$PS)
```

4.  Estimates of the causal contrasts are obtained as before

```{r tmle-estimates}
EY0pot.t <- mean(pop$ypred0.H)
EY1pot.t <- mean(pop$ypred1.H)
round(Contr(EY1pot.t, EY0pot.t), 4)
```

These are similar to previous results obtained by AIPW when computed for
the whole population with the correct exposure model, and they are very
close to the true values (e.g. RD=0.1743), too.
