---
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
knitr::opts_chunk$set(results = "hide", fig.show="hide", messages=FALSE, keep.source = TRUE, include = TRUE, eps = FALSE, prefix.string = "./graph/causInf2")
```

<!-- ```{r include=FALSE, warning=FALSE, message=FALSE} -->

<!-- base.pkg.list <- c("stats", "graphics", "grDevices", "datasets", "utils", "methods", "base") -->

<!-- loaded.pkg.list <- setdiff((.packages()), base.pkg.list) -->

<!-- for(pkg_ in loaded.pkg.list) { -->

<!--   detach(paste0("package:", pkg_), unload = TRUE, character.only = TRUE) -->

<!-- } -->

<!-- ``` -->

# Causal inference 2: Model-based estimation of causal estimands

Sources of inspiration: [Luque Fernandez, M.A. et al.
(2018)](https://doi.org/10.1002/sim.7628) *Stat Med*
2018;37(16):2530-2546 and\
[Smith et al. (2022)](https://doi.org/10.1002/sim.9234) *Stat Med*
2022;41(2):407-432.

## Introduction

We shall illustrate with simulated data the estimation of causal effects
of a binary exposure $X$ when the outcome $Y$ is also binary, and there
is a set of four covariates $Z = (Z_1, Z_2, Z_3, Z_4)$. As a background
story, we imagine a population of cancer patients, in whom the variables
and the assumed marginal distributions of the covariates are

| Variable | Description                                                           |
|:--------------|:-------------------------------------------------------|
| $X$      | treatment; 1: radiotherapy only, 0: radiotherapy + chemotherapy       |
| $Y$      | death during one year after diagnosis of cancer                       |
| $Z_1$    | sex; 0: man, 1: woman; $Z_1 \sim \text{Bern}(0.5)$                    |
| $Z_2$    | age group 0; `young`, 1: `old`; $Z_2 \sim \text{Bern}(0.65)$          |
| $Z_3$    | stage of cancer; 4 classes; $Z_3 \sim \text{DiscUnif}(1, \dots, 4)$   |
| $Z_4$    | comorbidity score; 5 classes; $Z_3 \sim \text{DiscUnif}(1, \dots, 5)$ |

For simplicity, covariates $Z_3$ and $Z_4$ are treated as continuous
variables in the models. The assumed causal diagram 
is drawn using `dagitty` and is shown below.

```{r dagitty, echo=TRUE, message=FALSE, warning=FALSE, results=TRUE, fig.show=TRUE}
library(dagitty)
d <-
  dagitty("dag {
  Z2 -> Z3 -> Y
  Z2 -> Z4 -> Y
  Z2 -> Y
  Z2 -> Z3 -> X
  Z2 -> Z4 -> X
  Z2 -> X
  Z1 -> Z3 -> Y
  Z1 -> Z4 -> Y
  Z1 -> Y
  Z1 -> Z3 -> X
  Z1 -> Z4 -> X
}")

dagitty::coordinates(d) <-
  list(
    x = c(X = 1, Y = 5, Z1 = 5, Z2 = 1, Z3 = 4, Z4 = 2),
    y = c(X = 2, Y = 2, Z1 = 0, Z2 = 0, Z3 = 1, Z4 = 1)
  )

plot(d)
```

For more generic notation, the probabilities of $Y=1$ will be expressed
as expectations, e.g. 
$$E(Y^{X=x}) = P(Y^{X=x}=1) \quad \text{and} \quad 
E(Y|X=x, Z=z) = P(Y=1|X=x, Z=z), $$
where $Z$ is the vector of relevant
covariates. 

The same principle is applied in expressing the conditional
probability of $X=1$, i.e. being exposed, given $Z=z$:
$$ E(X|Z=z) = P(X=1|Z=z). $$
The fitted or predicted probabilities
of $Y=1$ are denoted as fitted $\widehat{Y}$ or predicted values
$\widetilde{Y}$ of $Y$ with pertinent subscripts and/or superscripts.
-- Both $X$ and $Y$ are modelled by logistic regression. The expit-function
or inverse of the logit function is defined:
$\text{expit}(u) = 1/(1 + e^{-u})$, $u\in R$. This is equal to the
cumulative distribution function of the standard logistic distribution,
the values of which are returned in R by `plogis(u)`. The R function
that returns values of the logit-function is `qlogis()`.

The true model assumed for the dependence of exposure $X$ on covariates:
$$ E(X|Z_1 = z_1, \dots, Z_4 = z_4) =
      \text{expit}(-5 + 0.05z_2 + 0.25z_3 + 0.5z_4 + 0.4z_2z_4) . $$ The
assumed true model for the outcome is
$$ E(Y|X=x, Z_1 = z_1, \dots, Z_4 = z_4) =
           \text{expit}(-1 + x - 0.1z_1 + 0.35z_2 + 0.25z_3 +
                 0.20z_4 + 0.15z_2z_4) $$ 
Note that $X$ does not depend on $Z_1$, and that in both models
there is a product term $Z_2 Z_4$,
the effect of which appears weaker for the outcome model.

## Control of confounding

1.  Based on inspection of the causal diagram, can you provide
    justification for the claim that variables 
    $Z_2, Z_3$ and $Z_4$ form
    a proper subset of the four covariates, 
    which is sufficient to block all backdoor paths between $X$
    and $Y$ and thus remove confounding?

2.  Even though we have such a minimal sufficient set 
as indicated in item (a), why it still might be useful 
to include covariate
    $Z_1$, too, when modelling the outcome?

## Generation of target population and true models

1.  Load the necessary packages.

```{r packages 2, message=FALSE}
library(Epi)
library(stdReg)
library(PSweight)
library(SuperLearner)
library(tmle)
```

2.  Define two R-functions, which compute the expected values 
for the exposure and those for the outcome based on the assumed 
true exposure model and the true outcome model, respectively.

```{r true models}
EX <- function(z2, z3, z4) {
  plogis(-5 + 0.05 * z2 + 0.25 * z3 + 0.5 * z4 + 0.4 * z2 * z4)
}
EY <- function(x, z1, z2, z3, z4) {
  plogis(-1 + x - 0.1 * z1 + 0.35 * z2 + 0.25 * z3 +
    0.20 * z4 + 0.15 * z2 * z4)
}
```

3.  Define the function for the generation of data by 
simulating random values from pertinent probability
distributions based on the given    assumptions.

```{r data generation function}
genData <- function(N) {
  z1 <- rbinom(N, size = 1, prob = 0.5) # Bern(0.5)
  z2 <- rbinom(N, size = 1, prob = 0.65) # Bern(0.65)
  z3 <- trunc(runif(N, min = 1, max = 5), digits = 0) # DiscUnif(1,4)
  z4 <- trunc(runif(N, min = 1, max = 6), digits = 0) # DiscUnif(1,5)
  x <- rbinom(N, size = 1, prob = EX(z2, z3, z4))
  y <- rbinom(N, size = 1, prob = EY(x, z1, z2, z3, z4))
  data.frame(z1, z2, z3, z4, x, y)
}
```

4.  Generate a data frame `dd` for a big target population of 
500000  subjects

```{r popdata generation}
N <- 500000
set.seed(7777)
dd <- genData(N)
```

## Factual and counterfactual risks - associational and causal contrasts

1.  Compute the factual risks of death for the two exposure groups
$$ E(Y|X=x) = P(Y=1|X=x) = \frac{P(Y=1\ \&\ X=x)}{P(X=x)},
    \quad x=0,1, $$ 
in the whole target population, as well as their
    associational contrasts: risk difference, risk ratio, and odds
    ratio. Before that define a useful function

```{r Contr function and associational contrasts}
Contr <- function(mu1, mu0) {
  RD <- mu1 - mu0
  RR <- mu1 / mu0
  OR <- (mu1 / (1 - mu1)) / (mu0 / (1 - mu0))
  return(c(mu1, mu0, RD = RD, RR = RR, OR = OR))
}
Ey1fact <- with(dd, sum(y == 1 & x == 1) / sum(x == 1))
Ey0fact <- with(dd, sum(y == 1 & x == 0) / sum(x == 0))
round(Contr(Ey1fact, Ey0fact), 4)
```

How much bigger is the risk of death of those factually exposed to
radiotherapy only as compared with those receiving chemotherapy, too?

2.  Compute now the true **counterfactual** or 
  **potential risks**  of death
    $$ E(Y_i^{X_i=x}) = P(Y_i^{X_i=x}=1) = \pi_i^{X_i=x} $$ 
    for each
    individual under the alternative treatments or exposure values
    $x=0,1$ with given covariate values, then the 
    average or overall counterfactual risks $E(Y^{X=1}) = \pi^1$
    and $E(Y^{X=0}) = \pi^0$ in the population, 
    and finally the true **marginal causal contrasts**
    for the effect of $X$: 
    $$
    \begin{aligned}
     \text{RD} & = E(Y^{X=1})-E(Y^{X=0}), \qquad  \text{RR} = E(Y^{X=1})/E(Y^{X=0}), \\
     \text{OR} & = \frac{E(Y^{X=1})/[1 -  E(Y^{X=1})]}{E(Y^{X=0})/[1 -  E(Y^{X=0})] }
    \end{aligned}
    $$

```{r true contrasts}
dd <- transform(dd,
  EY1.ind = EY(x = 1, z1, z2, z3, z4),
  EY0.ind = EY(x = 0, z1, z2, z3, z4)
)
EY1pot <- mean(dd$EY1.ind)
EY0pot <- mean(dd$EY0.ind)
round(Contr(EY1pot, EY0pot), 4)
```

3.  Compare the associational contrasts computed in 
   item 4.1 with the
    causal contrasts in item 4.2. What do you conclude about
    confoundedness of the associational contrasts?

## Outcome modelling and estimation of causal contrasts by g-formula

As the first approach for estimating causal contrasts of interest
we apply the method of **standardization** or **g-formula**. 
It is based on a hopefully realistic enough model for 
$E(Y|X=x, Z=z)$, i.e. how the risk
of outcome is expected to depend on the exposure variable 
$X$ and on a
sufficient set $Z$ of confounders. The potential 
or counterfactual risks
$E(Y^{X=x}), x=0,1$, are marginal expectations of the above
quantities, standardized over the joint distribution of the confounders $Z$ in the
target population. 
$$ E(Y^{X=x}) = E_Z[E(Y|X=x,Z)]
       = \int E(Y|X=x, Z=z)dF_Z(z), \quad x=0,1. $$

1.  Assume now a *slightly misspecified* model `mY` 
 for the outcome, which contains only main effect terms
 of the explanatory variables:
  $$ \pi_i = E(Y_i|X_i=x_i, Z_{i1}=z_{i1}, \dots, Z_{i4}=z_{i4}) =
      \text{expit}\left(\beta_0 + \delta x_i +
      \sum_{j=1}^4 \beta_j z_{ij} \right) $$ 
 Fit this model on the
    target population using function `glm()`
    <!-- % in order to have an accurate -->
    <!-- % estimate of the possible bias due to misspecification of the outcome -->
    <!-- % model -->

```{r outcome model}
mY <- glm(y ~ x + z1 + z2 + z3 + z4, family = binomial, data = dd)
round(ci.lin(mY, Exp = TRUE)[, c(1, 5)], 3)
```

There is not much idea in looking at the standard errors or
confidence intervals in such a big population.

2.  For each subject $i$, compute the fitted individual risk
    $\widehat{Y_i}$ as well as the predicted potential
    (counterfactual) risks
    $\widetilde{Y_i}^{X_i=x}$ for both exposure levels $x=0,1$
    separately, keeping the individual values of the 
    $Z$-variables as they are.

```{r fitted risks and predicted potential risks}
dd$yh <- predict(mY, type = "response") #  fitted values
dd$yp1 <- predict(mY, newdata = data.frame(
  x = rep(1, N), # x=1
  dd[, c("z1", "z2", "z3", "z4")]
), type = "response")
dd$yp0 <- predict(mY, newdata = data.frame(
  x = rep(0, N), # x=0
  dd[, c("z1", "z2", "z3", "z4")]
), type = "response")
```

3.  Applying the method of **standardization** or 
**g-formula** compute now the
    point estimates 
    $$ \widehat{E}_g(Y^{X=x}) =
     \frac{1}{n} \sum_{i=1}^n \widetilde{Y}_i^{X_i=x}, \quad x=0,1. $$
    of the two counterfactual risks $E(Y^{X=1}) = \pi^1$ and
    $E(Y^{X=0})=\pi^0$ as well as  
    those of the marginal causal contrasts

```{r causal contrasts}
EY1pot.g <- mean(dd$yp1)
EY0pot.g <- mean(dd$yp0)
round(Contr(EY1pot.g, EY0pot.g), 4)
```

The marginal expectations $E_Z[E(X=x, Z)]$ taken over the 
joint distribution of
the confounders $Z$ are empirically estimated from 
the actual data representing the target population by simply
computing the arithmetic means of the individually
predicted values $\widetilde{Y_i}^{X_i=x}$ of the 
outcome for the two exposure levels.

Compare the estimated contrasts with the true ones in
item 4.2 above.
How big is the bias due to slight misspecification 
of the outcome model?
Compare in particular the estimate of the marginal 
OR here with the
conditional OR obtained in item 5.1 from the
pertinent coefficient in
the logistic model. Which one is closer to 1?

4.  Perform the same calculations using the tools 
in package `stdReg`
(see [Sjölander 2016](https://doi.org/10.1007/s10654-016-0157-3))

```{r stdReg}
mY.std <- stdGlm(fit = mY, data = dd, X = "x")
summary(mY.std)
round(summary(mY.std, contrast = "difference", reference = 0)$est.table, 4)
round(summary(mY.std, contrast = "ratio", reference = 0)$est.table, 4)
round(summary(mY.std,
  transform = "odds",
  contrast = "ratio", reference = 0
)$est.table, 4)
```

Check that you got the same point estimates as 
in the previous item.
Again, the confidence intervals are not very 
meaningful when analysing
the data covering the whole big target population. 
Of course, when
applied to real sample data they are relevant. 
In `stdReg` package, the
standard errors are obtained by the multivariate 
delta method built upon
M-estimation and robust sandwich estimator of the 
pertinent covariance
matrix, and approximate confidence intervals are derived 
from these in the usual way.



## Inverse probability weighting (IPW) by propensity scores


The next method is based on weighting each individual 
observation by the
inverse of the probability of belonging to that 
particular exposure group, which was realized, 
this probability being predicted by the
determinants of exposure.

1.  Fit first a somewhat misspecified 
model for the exposure including the main effects of the
    $Z$-variables only. 
    $$ 
    p_i = E(X_i| Z_{1i} = z_{1i}, \dots, Z_{4i} = z_{4i})
    = \text{expit}(\gamma_0 + \gamma_1 z_{1i} + \gamma_2 z_{2i} +
       \gamma_3 z_{i3} + \gamma_4 z_{4i} ), \quad i=1, \dots N 
    $$

```{r exposure model}
mX <- glm(x ~ z1 + z2 + z3 + z4,
  family = binomial(link = logit), data = dd
)
round(ci.lin(mX, Exp = TRUE)[, c(1, 5)], 4)
```

2.  Extract the **propensity scores**, i.e. 
fitted probabilities of
    belonging to exposure group 1: 
    PS$_i = \widehat{p}_i$, and
    compare their distribution between the two groups.

```{r propScore}
dd$PS <- predict(mX, type = "response")
summary(dd$PS)
with(subset(dd, x == 0), plot(density(PS), lty = 2))
with(subset(dd, x == 1), lines(density(PS), lty = 1))
```

How different are the distributions? Are they sufficiently overlapping?

3.  Compute the weights
$$
\begin{aligned}
 W_i & = \frac{1}{\text{PS}_i}, \quad \text{when }\ X_i=1, \\
 W_i & = \frac{1}{1-\text{PS}_i}, \quad \text{when }\ X_i=0 . 
\end{aligned} 
$$
 Look at the sum as well as
    the distribution summary of the weights in the exposure groups. The
    sum of weights should be close to $N$ in both groups.

```{r weights}
dd$w <- ifelse(dd$x == 1, 1 / dd$PS, 1 / (1 - dd$PS))
with(dd, tapply(w, x, sum))
```

4.  Compute now the weighted estimates of the 
potential or counterfactual risks for
    both exposure categories 
    $$ \widehat{E}_w(Y^{X = x}) =
    \frac{ \sum_{i=1}^n {\mathbf 1}_{ \{X_i=x\} } W_i Y_i }
       {\sum_{i=1}^n {\mathbf 1}_{ \{X_i=x\} }W_i} =
     \frac{ \sum_{X_i = x} W_i Y_i }{\sum_{X_i=x} W_i}, \quad x = 0,1, $$
    and their causal contrasts, for instance
    $$ \widehat{\text{RD}}_{w} = \widehat{E}_w(Y^{X = 1}) -
                  \widehat{E}_w(Y^{X = 0})
  =  \frac{ \sum_{i=1}^n X_i W_i Y_i }{\sum_{i=1}^n X_i W_i} -
  \frac{ \sum_{i=1}^n (1-X_i) W_i Y_i }{\sum_{i=1}^n (1-X_i) W_i}
    $$

```{r ipw-estimate}
EY1pot.w <- sum(dd$x * dd$w * dd$y) / sum(dd$x * dd$w)
EY0pot.w <- sum((1 - dd$x) * dd$w * dd$y) / sum((1 - dd$x) * dd$w)
round(Contr(EY1pot.w, EY0pot.w), 4)
```

These estimates seem to be somewhat downward biased when 
comparing to
true values. Could this be because of omitting 
the relatively strong
product term effect of $Z_2$ and $Z_4$?


## Improving IPW estimation and using R package `PSweight`

We now try to improve IPW-estimation by a richer exposure 
model. In
computations we shall utilize the R package 
`PSweight` (see [PSweight
vignette](https://cran.r-project.org/web/packages/PSweight/vignettes/vignette.pdf)).

1.  First, we compute the propensity scores
and weights from a more flexible exposure model,
    which contains all pairwise product terms 
    of the parents of $X$.
    According to the causal diagram, $Z_1$ is 
    not in that subset, so it
    is left out. The exposure model is 
    specified and the weights are
    obtained as follows using function
    `SumStat()`:

```{r PSweight, fig=FALSE}
mX2 <- glm(x ~ (z2 + z3 + z4)^2, family = binomial, data = dd)
round(ci.lin(mX2, Exp = TRUE)[, c(1, 5)], 3)
psw2 <- SumStat(
  ps.formula = mX2$formula, data = dd,
  weight = c("IPW", "treated", "overlap")
)
dd$PS2 <- psw2$propensity[, 2] 
dd$w2 <- ifelse(dd$x == 1, 1 / dd$PS2, 1 / (1 - dd$PS2)) 
plot(density(dd$PS2[dd$x == 0]), lty = 2)
lines(density(dd$PS2[dd$x == 1]), lty = 1)
```

Note that apart from ordinary IPW, other types of 
weights can also also
obtained. These are relevant when estimating other 
kinds of causal
contrasts, like "average treatment effect among the treated" 
(ATT, see below) and "average treatment effect in the overlap 
(or equipoise) population" (ATO).

2.  `PSweight` includes some useful tools to examine the
properties of the distribution and to check the balance 
of the propensity scores, for instance

```{r check balance, fig=FALSE}
plot(psw2, type = "balance", metric = "PSD")
```

It is desirable that the horisontal values of 
these measures for given weights are less than 0.1.

3.  Estimation and reporting of the causal contrasts. For relative
    contrasts, the summary method provides the results 
    on the log-scale; therefore $\exp$-transformation

```{r ipw-estimation}
ipw2est <- PSweight(ps.formula = mX2, yname = "y", data = dd, weight = "IPW")
ipw2est
summary(ipw2est)
(logRR.ipw2 <- summary(ipw2est, type = "RR"))
round(exp(logRR.ipw2$estimates[c(1, 4, 5)]), 3)
round(exp(summary(ipw2est, type = "OR")$estimates[c(1, 4, 5)]), 3)
```

Compare these with the previous IPW estimate  as
well as the true values. Have we obtained nearly unbiased results?

The standard errors provided by `PSweight` 
are by default based on the
empirical sandwich covariance matrix and application 
of delta method as
appropriate. Bootstrapping is also possible but is 
computationally very
intensive and is recommended to be used only in relatively small
samples.

## Effect of exposure among those exposed 

If we are interested in the 
causal contrasts describing the **effect
    of exposure among those exposed** (like ATE), 
    the relevant factual and
    counterfactual risks in that subset are

$$
\begin{aligned}
 \pi^1_1 & = E(Y^{X=1}|X=1) = E(Y|X=1) = \pi_1, \\
 \pi^0_1 & = E(Y^{X=0}|X=1) = \sum_{X_i=1} E(Y|X=0, Z=z)P(Z=z|X=1)
\end{aligned}
$$

We are thus making and "observed vs. expected" comparison, 
in which the
$z$-specific risks in the unexposed are weighted by the
distribution of
$Z$ in the exposed subset of the target population. 
The risks and their
contrasts are estimated from the fit of the outcome model:

```{r g-formula-att}
EY1att.g <- mean(subset(dd, x == 1)$yp1)
EY0att.g <- mean(subset(dd, x == 1)$yp0)
round(Contr(EY1att.g, EY0att.g), 4)
```

Compare the results here with those for the whole target population. What do you observe? 


2. Have you any guess about the causal effect of exposure among
the unexposed; is it bigger or smaller than among 
the exposed or among the whole population?

3.  Incidentally, the true causal contrasts among the 
exposed based on the true model are similarly obtained 
from the quantities in item 4.2 above:

```{r true among exposed}
EY1att <- mean(subset(dd, x == 1)$EY1.ind)
EY0att <- mean(subset(dd, x == 1)$EY0.ind)
round(Contr(EY1att, EY0att), 4)
```

Compare the estimates in the previous item with the 
true values obtained here.

4.  When wishing to estimate 
the effect of exposure among the exposed using the IPW
approach, then the weights are $W_i = 1$ for the exposed and
    $W_i = \text{PS}_i/(1-\text{PS}_i)$ for the unexposed.
    Call again `PSweight` but with another choice of weight:

```{r ps-estimation-att}
psatt <- PSweight(ps.formula = mX2, yname = "y", data = dd, weight = "treated")
psatt
round(summary(psatt)$estimates[1], 4)
round(exp(summary(psatt, type = "RR")$estimates[1]), 3)
round(exp(summary(psatt, type = "OR")$estimates[1]), 3)
```

Compare the results here with those obtained by 
g-formula in item 8.1
and with the true contrasts above.

## Double robust estimation by augmented IPW

Let us attempt to correct the estimates by a **double 
robust** (DR) approach
called **augmented IPW estimation** (AIPW), which 
combines the g-formula and
the IPW approach. The classical 
AIPW-estimator can be expressed in two ways:
either an IPW-corrected g-formula estimator, or a g-corrected
IPW-estimator.

$$
\begin{aligned}
 \widehat{E}_a(Y^{X=x}) 
       & = \widehat{E}_g(Y^{X=x}) +
\frac{1}{n} \sum_{i=1}^n  {\mathbf 1}_{\{X_i=x\}} W_i 
                   ( Y_i - \widetilde{Y}_i^{X_i=x} )   \\
       & = \widehat{E}_w(Y^{X=x}) +
\frac{1}{n} \sum_{i=1}^n ( 1 - {\mathbf 1}_{\{X_i=x\}} W_i ) 
                      \widetilde{Y}_i^{X_i=x}.
\end{aligned}
$$

1. We shall first combine the results from the slightly
misspecified outcome model with those from the more 
misspecified exposure model.

```{r aipw1}
EY1pot.a <- EY1pot.g + mean( 1*(dd$x==1) * dd$w * (dd$y - dd$yp1) )
EY0pot.a <- EY0pot.g + mean( 1*(dd$x==0) * dd$w * (dd$y - dd$yp0) )
round(Contr(EY1pot.a, EY0pot.a), 4)
```

Compare these results with those obtained by g-formula and by
non-augmented IPW method. Was augmentation successful?

2. Let us the look, how close we get when combining the results
from the slightly misspecified outcome model with the correct
exposure model using the alternative AIPW-formula

```{r aipw2}
EY1pot.w2 <- ipw2est$muhat[2]
EY0pot.w2 <- ipw2est$muhat[1]
EY1pot.a2 <- EY1pot.w2 + mean( (1 - 1*(dd$x==1) * dd$w2) * dd$yp1 )
EY0pot.a2 <- EY0pot.w2 + mean( (1 - 1*(dd$x==0) * dd$w2) * dd$yp0 )
round(Contr(EY1pot.a2, EY0pot.a2), 4)
```

Compare the results with previous ones.
How successful was augmentation now?

AIPW-estimates and confidence
intervals for the causal contrasts of interest can
be obtained, for instance, using `PSweight` by adding
the model formula of the outcome model
as the value for the argument `out.formula`.

## Targeted maximum likelihood estimation (TMLE)

We now consider now another double robust approach, 
known as **targeted
maximum likelihood estimation** (TMLE). 
It also corrects the estimator
obtained from the outcome model by elements that are derived from the exposure model. The corrections are, though, not as
intuitive as those in AIPW. See [Schuler and Rose
(2017)](https://doi.org/10.1093/aje/kww165) for more details.

1.  The first step is to utilize the propensity scores 
obtained above for the correct exposure model 
and define the "clever covariates"

```{r clever covariates}
dd$H1 <- dd$x / dd$PS2
dd$H0 <- (1 - dd$x) / (1 - dd$PS2)
```

2.  Then, a working model is fitted for the outcome, 
in which the clever
    covariates are explanatory variables, but the model 
    also includes
    the previously fitted linear predictor
    $\widehat{\eta}_i = \text{logit}(\widehat Y_i)$ 
    from the original
    outcome model `mY` as an offset term; see item 5.2.
    Moreover, the intercept is removed.

```{r model with clever covariates}
epsmod <- glm(y ~ -1 + H0 + H1 + offset(qlogis(yh)),
  family = binomial(link = logit), data = dd
)
eps <- coef(epsmod)
eps
```

3.  The logit-transformed predicted values $\widetilde{Y}_i^{X_i=1}$ and
    $\widetilde{Y}_i^{X_i=0}$ of 
    the potential or counterfactual individual risks from
    the original outcome model are now corrected by the estimated
    coefficients of the clever covariates, and the corrected
    predictions are returned to the original scale.

```{r tmle predictions}
ypred0.H <- plogis(qlogis(dd$yp0) + eps[1] / (1 - dd$PS2))
ypred1.H <- plogis(qlogis(dd$yp1) + eps[2] / dd$PS2)
```

4. Estimates of the causal contrasts:

```{r tmle-estimates}
EY0pot.t <- mean(ypred0.H)
EY1pot.t <- mean(ypred1.H)
round(Contr(EY1pot.t, EY0pot.t), 4)
```

Compare these with previous results and with the true values.

## Double robust estimation with `AIPW` and `tmle` packages

It may be difficult to specify 
conventional generalized linear models 
or even generalized additive 
models for exposure and outcome which are 
sufficiently realistic, yet which do not suffer from overfitting.
 Modern approaches of **statistical learning**, aka
"machine learning" provide tools for flexible modelling,
which may be used to reduce the risk of misspecification,
 if thoughtfully applied. 

There are a few R packages in which some general
algorithmic approaches for supervised learning
are implemented for estimating causal parameters.
For instance, package `AIPW` 
[Zhong et al. (2021)](https://doi.org/10.1093/aje/kwab207)
utilizes 
several learning algorithms for exposure and outcome modelling
and then performs AIPW estimation  of the
parameters of interest coupled with 
calculation of confidence intervals. Package `tmle`
[Karim and Frank (2021)](https://ehsanx.github.io/TMLEworkshop/)
performs same tasks but uses the TMLE approach 
in estimation.

Both `AIPW` and `tmle` lean on the
 `SuperLearner` package, which uses multiple learning 
 algorithms (e.g. GLM, GAM, Random Forest, Recursive Partitioning,
 Gradient Boosting, etc.) for constructing
 predictions of the counterfactual quantities, and 
 then creates an optimal weighted average of those models, 
 aka an "ensemble". These algorithms are computationally
 highly intensive. Fitting models with only 3 or 4
 covariates as in this practical to a data set of  
 observational units on our target cohort of 500,000 subjects
 would take hours on an ordinary laptop. With a study population
 of a couple of thousands it takes several minutes. 



