---
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include=FALSE}
knitr::opts_chunk$set(results = "hide", keep.source = TRUE, include = TRUE, eps = FALSE, prefix.string = "./graph/causInf2")
```

<!-- ```{r include=FALSE, warning=FALSE, message=FALSE} -->

<!-- base.pkg.list <- c("stats", "graphics", "grDevices", "datasets", "utils", "methods", "base") -->

<!-- loaded.pkg.list <- setdiff((.packages()), base.pkg.list) -->

<!-- for(pkg_ in loaded.pkg.list) { -->

<!--   detach(paste0("package:", pkg_), unload = TRUE, character.only = TRUE) -->

<!-- } -->

<!-- ``` -->

# Causal inference 2: Model-based estimation of causal estimands

Sources of inspiration: [Luque Fernandez, M.A. et al.
(2018)](https://doi.org/10.1002/sim.7628) *Stat Med*
2018;37(16):2530-2546 and\
[Smith et al. (2022)](https://doi.org/10.1002/sim.9234) *Stat Med*
2022;41(2):407-432.

We shall illustrate with simulated data the estimation of causal effects
of a binary exposure $X$ when the outcome $Y$ is also binary, and there
is a set of four covariates $Z = (Z_1, Z_2, Z_3, Z_4)$. As a background
story, we imagine a population of cancer patients, in whom the variables
and the assumed marginal distributions of the covariates are

| Variable | Description                                                           |
|:--------------|:-------------------------------------------------------|
| $X$      | treatment; 1: radiotherapy only, 0: radiotherapy + chemotherapy       |
| $Y$      | death during one year after diagnosis of cancer                       |
| $Z_1$    | sex; 0: man, 1: woman; $Z_1 \sim \text{Bern}(0.5)$                    |
| $Z_2$    | age group 0; `young`, 1: `old`; $Z_2 \sim \text{Bern}(0.65)$          |
| $Z_3$    | stage of cancer; 4 classes; $Z_3 \sim \text{DiscUnif}(1, \dots, 4)$   |
| $Z_4$    | comorbidity score; 5 classes; $Z_3 \sim \text{DiscUnif}(1, \dots, 5)$ |

For simplicity, covariates $Z_3$ and $Z_4$ are treated as continuous
variables in the models. The assumed causal diagram is shown below.

```{r echo=FALSE, message=FALSE, warning=FALSE, results=TRUE, fig.show=TRUE}
library(dagitty)
d <-
  dagitty("dag {
  Z2 -> Z3 -> Y
  Z2 -> Z4 -> Y
  Z2 -> Y
  Z2 -> Z3 -> X
  Z2 -> Z4 -> X
  Z2 -> X
  Z1 -> Z3 -> Y
  Z1 -> Z4 -> Y
  Z1 -> Y
  Z1 -> Z3 -> X
  Z1 -> Z4 -> X
}")

dagitty::coordinates(d) <-
  list(
    x = c(X = 1, Y = 5, Z1 = 5, Z2 = 1, Z3 = 4, Z4 = 2),
    y = c(X = 3, Y = 3, Z1 = 1, Z2 = 1, Z3 = 2, Z4 = 2)
  )

plot(d)
```

For more generic notation, the probabilities of $Y=1$ will be expressed
as expectations, e.g. $E(Y^{X=x}) = P(Y^{X=x}=1)$ and
$E(Y|X=x, Z=z) = P(Y=1|X=x, Z=z)$, where $Z$ is the vector of relevant
covariates. The same principle is applied in expressing the conditional
probability of $X=1$ given $Z=z$. The fitted or predicted probabilities
of $Y=1$ are denoted as fitted $\widehat{Y}$ or predicted values
$\widetilde{Y}$ of $Y$ with pertinent subscripts and/or superscripts.
Both $X$ and $Y$ are modelled by logistic regression. The expit-function
or inverse of the logit function is defined:
$\text{expit}(u) = 1/(1 + e^{-u})$, $u\in R$. This is equal to the
cumulative distribution function of the standard logistic distribution,
the values of which are returned in R by `plogis(u)`. The R function
that returns values of the logit-function is `qlogis()`.

The true model assumed for the dependence of exposure $X$ on covariates:
$$ E(X|Z_1 = z_1, \dots, Z_4 = z_4) =
      \text{expit}(-5 + 0.05z_2 + 0.25z_3 + 0.5z_4 + 0.4z_2z_4) . $$ The
assumed true model for the outcome is
$$ E(Y|X=x, Z_1 = z_1, \dots, Z_4 = z_4) =
           \text{expit}(-1 + x - 0.1z_1 + 0.35z_2 + 0.25z_3 +
                 0.20z_4 + 0.15z_2z_4) $$ Note that $X$ does not depend
on $Z_1$, and that in both models there is a product term $Z_2 Z_4$,
which appears weaker for the outcome model.

## Control of confounding

1.  Based on inspection of the causal diagram, can you provide
    justification for the claim that variables $Z_2, Z_3$ and $Z_4$ form
    a proper subset of the four covariates, which is sufficient to block
    all backdoor paths between $X$ and $Y$ and thus remove confounding?

2.  Even though we have such a minimal sufficient set as indicated in
    item (a), why could it still be worth while to include covariate
    $Z_1$, too, when modelling the outcome?

## Generation of target population and true models

1.  Load the necessary packages.

```{r packages 2}
library(Epi)
library(stdReg)
library(PSweight)
library(SuperLearner)
library(tmle)
```

2.  Define the R-functions which computes expected values for the
    exposure and outcome based on the assumed true outcome model and the
    true exposure model.

```{r true models}
EX <- function(z2, z3, z4) {
  plogis(-5 + 0.05 * z2 + 0.25 * z3 + 0.5 * z4 + 0.4 * z2 * z4)
}
EY <- function(x, z1, z2, z3, z4) {
  plogis(-1 + x - 0.1 * z1 + 0.35 * z2 + 0.25 * z3 +
    0.20 * z4 + 0.15 * z2 * z4)
}
```

3.  Define the function for the generation of data by simulating random
    values from pertinent probability distributions based on the given
    assumptions.

```{r data generation function}
genData <- function(N) {
  z1 <- rbinom(N, size = 1, prob = 0.5) # Bern(0.5)
  z2 <- rbinom(N, size = 1, prob = 0.65) # Bern(0.65)
  z3 <- trunc(runif(N, min = 1, max = 5), digits = 0) # DiscUnif(1,4)
  z4 <- trunc(runif(N, min = 1, max = 6), digits = 0) # DiscUnif(1,5)
  x <- rbinom(N, size = 1, prob = EX(z2, z3, z4))
  y <- rbinom(N, size = 1, prob = EY(x, z1, z2, z3, z4))
  data.frame(z1, z2, z3, z4, x, y)
}
```

4.  Generate a data frame `dd` for a big target population of 500000
    subjects

```{r popdata generation}
N <- 500000
set.seed(7777)
dd <- genData(N)
```

## Factual and counterfactual risks - associational and causal contrasts

1.  Compute the factual risks of death for the two exposure groups
    $$ E(Y|X=x) = P(Y=1|X=x) = \frac{P(Y=1\ \&\ X=x)}{P(X=x)},
    \quad x=0,1, $$ in the whole target population, as well as their
    associational contrasts: risk difference, risk ratio, and odds
    ratio. Before that define a useful function

```{r association}
Contr <- function(mu1, mu0) {
  RD <- mu1 - mu0
  RR <- mu1 / mu0
  OR <- (mu1 / (1 - mu1)) / (mu0 / (1 - mu0))
  return(c(mu1, mu0, RD = RD, RR = RR, OR = OR))
}
Ey1 <- with(dd, sum(y == 1 & x == 1) / sum(x == 1))
Ey0 <- with(dd, sum(y == 1 & x == 0) / sum(x == 0))
round(Contr(Ey1, Ey0), 4)
```

How much bigger is the risk of death of those factually exposed to
radiotherapy only as compared with those receiving chemotherapy, too?

2.  Compute now first the counterfactual risks of death
    $E(Y_i^{X_i=x}) = P(Y_i^{X_i=x}=1) = \pi_i^{X_i=x}$ for each
    individual under the alternative treatments or exposure values
    $x=0,1$ with given covariate values, then the average or overall
    counterfactual risks $E(Y^{X=1}) = \pi^1$ and $E(Y^{X=0}) = \pi^0$
    in the population, and finally the true marginal causal contrasts
    for the effect of $X$: $$
    \begin{aligned}
     \text{RD} & = E(Y^{X=1})-E(Y^{X=0}), \qquad  \text{RR} = E(Y^{X=1})/E(Y^{X=0}), \\
     \text{OR} & = \frac{E(Y^{X=1})/[1 -  E(Y^{X=1})]}{E(Y^{X=0})/[1 -  E(Y^{X=0})] }
    \end{aligned}
    $$

```{r true contrasts}
dd <- transform(dd,
  EY1.ind = EY(x = 1, z1, z2, z3, z4),
  EY0.ind = EY(x = 0, z1, z2, z3, z4)
)
EY1 <- mean(dd$EY1.ind)
EY0 <- mean(dd$EY0.ind)
round(Contr(EY1, EY0), 4)
```

3.  Compare the associational contrasts computed in in item (a) with the
    causal contrasts in item (b). What do you conclude about
    confoundedness of the associational contrasts?

## Outcome modelling and estimation of causal contrasts by g-formula

As the first approach for estimating causal contrasts of interest we
apply the method of standardization or g-formula. It is based on a
hopefully realistic enough model for $E(Y|X=x, Z=z)$, i.e. how the risk
of outcome is expected to depend on the exposure variable $X$ and on a
sufficient set $Z$ of confounders. The counterfactual risks
$E(Y^{X=x}), x=0,1$, are marginal expectations of the above quantities,
standardized over the joint distribution of the confounders $Z$ in the
target population. $$ E(Y^{X=x}) = E_Z[E(Y|X=x,Z)]
       = \int E(Y|X=x, Z=z)dF_Z(z), \quad x=0,1. $$

1.  Assume now a *slightly misspecified* model `mY` for the outcome,
    which contains only main effect terms of the explanatory variables:
    $$ \pi_i = E(Y_i|X_i=x_i, Z_{i1}=z_{i1}, \dots, Z_{i4}=z_{i4}) =
      \text{expit}\left(\beta_0 + \delta x_i +
      \sum_{j=1}^4 \beta_j z_{ij} \right) $$ Fit this model on the
    target population using function `glm()`
    <!-- % in order to have an accurate -->
    <!-- % estimate of the possible bias due to misspecification of the outcome -->
    <!-- % model -->

```{r outcome model}
mY <- glm(y ~ x + z1 + z2 + z3 + z4, family = binomial, data = dd)
round(ci.lin(mY, Exp = TRUE)[, c(1, 5)], 3)
```

There is not much idea in looking at the standard errors or confidence
intervals in such a big population.

2.  For each subject $i$, compute the fitted individual risk
    $\widehat{Y_i}$ as well as the predicted counterfactual risks
    $\widetilde{Y_i}^{X_i=x}$ for both exposure levels $x=0,1$
    separately, keeping the individual values of the $Z$-variables as
    they are.

```{r predict}
dd$yh <- predict(mY, type = "response") #  fitted values
dd$yp1 <- predict(mY, newdata = data.frame(
  x = rep(1, N), # x=1
  dd[, c("z1", "z2", "z3", "z4")]
), type = "response")
dd$yp0 <- predict(mY, newdata = data.frame(
  x = rep(0, N), # x=0
  dd[, c("z1", "z2", "z3", "z4")]
), type = "response")
```

3.  Applying the method of standardization or g-formula compute now the
    point estimates $$ \widehat{E}_g(Y^{X=x}) =
     \frac{1}{n} \sum_{i=1}^n \widetilde{Y}_i^{X_i=x}, \quad x=0,1. $$
    of the two counterfactual risks $E(Y^{X=1}) = \pi^1$ and
    $E(Y^{X=0})=\pi^0$ as well as the marginal causal contrasts

```{r causal contrasts}
EY1.g <- mean(dd$yp1)
EY0.g <- mean(dd$yp0)
round(Contr(EY1.g, EY0.g), 4)
```

The expectations $E_Z[E(X=x, Z)]$ taken over the joint distribution of
the confounders $Z$ are empirically estimated from the data by simply
computing the arithmetic means of the individually predicted values
$\widetilde{Y_i}^{X_i=x}$ of the outcome for the two exposure levels.

Compare the estimated contrasts with the true ones in item 3(b) above.
How big is the bias due to slight misspecification of the outcome model?
Compare in particular the estimate of the marginal OR here with the
conditional OR obtained in item (a) from the pertinent coefficient in
the logistic model. Which one is closer to 1?

4.  Perform the same calculations using the tools in package `stdReg`
    (see [Sjölander 2016](https://doi.org/10.1007/s10654-016-0157-3))

```{r stdReg}
mY.std <- stdGlm(fit = mY, data = dd, X = "x")
summary(mY.std)
round(summary(mY.std, contrast = "difference", reference = 0)$est.table, 4)
round(summary(mY.std, contrast = "ratio", reference = 0)$est.table, 4)
round(summary(mY.std,
  transform = "odds",
  contrast = "ratio", reference = 0
)$est.table, 4)
```

Check that you got the same point estimates as in the previous item.
Again, the confidence intervals are not very meaningful when analysing
the data covering the whole big target population. Of course, when
applied to real sample data they are relevant. In `stdReg` package, the
standard errors are obtained by the multivariate delta method built upon
M-estimation and robust sandwich estimator of the pertinent covariance
matrix, and approximate confidence intervals are derived from these in
the usual way.

5.  If we are interested in the causal contrasts describing the effect
    of exposure among those exposed (like ATT), the relevant factual and
    counterfactual risks in that subset are

$$
\begin{aligned}
 \pi^1_1 & = E(Y^{X=1}|X=1) = E(Y|X=1) = \pi_1, \\
 \pi^0_1 & = E(Y^{X=0}|X=1) = \sum_{X_i=1} E(Y|X=0, Z=z)P(Z=z|X=1)
\end{aligned}
$$

We are thus making and *observed vs. expected* comparison, in which the
$z$-specific risks in the unexposed are weighted by the distribution of
$Z$ in the exposed subset of the target population. The risks and their
contrasts are estimated from the fit of the outcome model:

```{r g-formula-att}
EY1att.g <- mean(subset(dd, x == 1)$yp1)
EY0att.g <- mean(subset(dd, x == 1)$yp0)
round(Contr(EY1att.g, EY0att.g), 4)
```

Compare the results here with those for the whole target population.
What do you observe? Any guess about the causal effect of exposure among
the unexposed; is it bigger or smaller than among the exposed or among
the whole population?

1.  Incidentally, the true causal contrasts among the exposed based on
    the true model are similarly obtained from the quantities in item
    3(b) above:

```{r true among exposed}
EY1att <- mean(subset(dd, x == 1)$EY1.ind)
EY0att <- mean(subset(dd, x == 1)$EY0.ind)
round(Contr(EY1att, EY0att), 4)
```

Compare the estimates in the previous item with the true values obtained
here.

## Inverse probability weighting (IPW) by

```         
propensity scores, and augmented IPW
```

The next method is based on weighting each individual observation by the
inverse of the probability of belonging to that particular exposure
group, which was realized, this probability being predicted by
determinants of exposure.

1.  Fit first a model for the exposure including main effects of the
    $Z$-variables only. $$ 
    p_i = E(X_i| Z_{1i} = z_{1i}, \dots, Z_{4i} = z_{4i})
    = \text{expit}(\gamma_0 + \gamma_1 z_{1i} + \gamma_2 z_{2i} +
       \gamma_3 z_{i3} + \gamma_4 z_{4i} ), \quad i=1, \dots N 
    $$

```{r exposure model}
mX <- glm(x ~ z1 + z2 + z3 + z4,
  family = binomial(link = logit), data = dd
)
round(ci.lin(mX, Exp = TRUE)[, c(1, 5)], 4)
```

2.  Extract the propensity scores, i.e. fitted probabilities of
    belonging to exposure group 1: $$ PS_i = \widehat{p_i} $$, and
    compare their distribution between the two groups.

```{r propScore, fig=FALSE}
dd$PS <- predict(mX, type = "response")
summary(dd$PS)
with(subset(dd, x == 0), plot(density(PS), lty = 2))
with(subset(dd, x == 1), lines(density(PS), lty = 1))
```

How different are the distributions? Are they sufficiently overlapping?

3.  Compute the weights $W_i = 1/\text{PS}_i$, when $X_i=1$, and
    $W_i = 1/(1-\text{PS}_i)$, when $X_i=0$. Look at the sum as well as
    the distribution summary of the weights in the exposure groups. The
    sum of weights should be close to $n$ in both groups.

```{r weights}
dd$w <- ifelse(dd$x == 1, 1 / dd$PS, 1 / (1 - dd$PS))
with(dd, tapply(w, x, sum))
```

4.  Compute now the weighted estimates of the counterfactual risks for
    both exposure categories $$ \widehat{E}_w(Y^{X = x}) =
    \frac{ \sum_{i=1}^n {\mathbf 1}_{ \{X_i=x\} } W_i Y_i }
       {\sum_{i=1}^n {\mathbf 1}_{ \{X_i=x\} }W_i} =
     \frac{ \sum_{X_i = x} W_i Y_i }{\sum_{X_i=x} W_i}, \quad x = 0,1, $$
    and their causal contrasts, for instance
    $$ \widehat{\text{RD}}_{w} = \widehat{E}_w(Y^{X = 1}) -
                  \widehat{E}_w(Y^{X = 0})
     =   \frac{ \sum_{i=1}^n X_i W_i Y_i }{\sum_{i=1}^n X_i W_i} -
        \frac{ \sum_{i=1}^n (1-X_i) W_i Y_i }{\sum_{i=1}^n (1-X_i) W_i}
    $$

```{r ipw-estimate}
EY1.w <- sum(dd$x * dd$w * dd$y) / sum(dd$x * dd$w)
EY0.w <- sum((1 - dd$x) * dd$w * dd$y) / sum((1 - dd$x) * dd$w)
round(Contr(EY1.w, EY0.w), 4)
```

These estimates seem to be somewhat downward biased when comparing to
true values. Could this be because of omitting the relatively strong
product term effect of $Z_2$ and $Z_4$?

5.Let us attempt to correct the estimates by a double robust approach
called augmented IPW estimation (AIPW), which combines the g-formula and
the IPW approach. The AIPW-estimator can be expressed in two ways:
either an IPW-corrected g-formula estimator, or a g-corrected
IPW-estimator.

$$
\begin{aligned}
 \widehat{E}_a(Y^{X=x}) & = \widehat{E}_g(Y^{X=x}) +
   \frac{1}{n} \sum_{i=1}^n \frac{ {\mathbf 1}_{\{X_i=x\}} W_i ( Y_i - \widetilde{Y}_i^{X_i=x} ) }
   {\sum_{i=1}^n {\mathbf 1}_{\{X_i=x\}} W_i} \\
       & =   \widehat{E}_w(Y^{X=x}) -
    \frac{1}{n} \sum_{i=1}^n \left[ \frac{ {\mathbf 1}_{\{X_i=x\}} W_i }
            {\sum_{i=1}^n {\mathbf 1}_{\{X_i=x\}} W_i } - 1 \right] \widetilde{Y}_i^{X_i=x}.
\end{aligned}
$$

```{r aipw}
EY1.a <- EY1.g + mean(dd$x * (dd$y - dd$yp1) * dd$w / sum(dd$x * dd$w))
##  or   EY1.w - mean( ( ( dd$x*dd$w /sum(dd$x*dd$w) ) - 1 )*dd$yp1 )
EY0.a <- EY0.g + mean((1 - dd$x) * (dd$y - dd$yp0) * dd$w / sum((1 - dd$x) * dd$w))
##  or   EY0.w - mean( ( ( (1-dd$x)*dd$w/sum((1-dd$x)*dd$w) ) - 1 )*dd$yp0 )
round(Contr(EY1.a, EY0.a), 4)
```

Compare these results with those obtained by g-formula and by
non-augmented IPW method. Was augmentation successful?

## Improving IPW estimation and using R package `PSweight`

We now try to improve IPW-estimation by a richer exposure model. In
computations we shall utilize the R package `PSweight` (see [PSweight
vignette](https://cran.r-project.org/web/packages/PSweight/vignettes/vignette.pdf)).

1.  First, we compute the weights from a more flexible exposure model
    which contains all pairwise product terms of the parents of $X$.
    According to the causal diagram, $Z_1$ is not in that subset, so it
    is left out. The exposure model is specified and the weights are
    obtained as follows.

```{r PSweight, fig=FALSE}
mX2 <- glm(x ~ (z2 + z3 + z4)^2, family = binomial, data = dd)
round(ci.lin(mX2, Exp = TRUE)[, c(1, 5)], 3)
psw <- SumStat(
  ps.formula = mX2$formula, data = dd,
  weight = c("IPW", "treated", "overlap")
)
dd$PS2 <- psw$propensity[, 2] # propensity scores extracted
plot(density(dd$PS2[dd$x == 0]), lty = 2)
lines(density(dd$PS2[dd$x == 1]), lty = 1)
```

Note that apart from ordinary IPW, other types of weights can also also
obtained. These are relevant when estimating other kinds of causal
contrasts, like *average treatment effect among the treated* (ATT) and
\`\`average treatment effect in the overlap (or equipoise) population''
(ATO).

2.  `PSweight` includes some useful tools to examine the properties of
    the distribution and to check the balance of the propensity scores,
    for instance

```{r check balance, fig=FALSE}
plot(psw, type = "balance", metric = "PSD")
```

It is desirable that the horisontal values of these measures for given
weights are less than 0.1.

3.  Estimation and reporting of the causal contrasts. For relative
    contrasts, the summary method provides the results on the log-scale.

```{r ipw-estimation}
ipwest <- PSweight(ps.formula = mX2, yname = "y", data = dd, weight = "IPW")
ipwest
summary(ipwest)
(logRR.ipw <- summary(ipwest, type = "RR"))
round(exp(logRR.ipw$estimates[c(1, 4, 5)]), 3)
round(exp(summary(ipwest, type = "OR")$estimates[c(1, 4, 5)]), 3)
```

Compare these with the previous IPW estimate and the AIPW estimate as
well as the true values. Have we obtained nearly unbiased results?

The standard errors provided by `PSweight` are by default based on the
empirical sandwich covariance matrix and application of delta method as
appropriate. Bootstrapping is also possible but is computationally very
intensive and is recommended to be used only in relatively small
samples.

4.  If we are interested in the effect of exposure among the exposed
    (like ATT) then the weights are $W_i = 1$ for the exposed and
    $W_i = \text{PS}_i/(1-\text{PS}_i)$ for the unexposed. Call again
    `PSweight` but with another choice of weight:

```{r ps-estimation-att}
psatt <- PSweight(ps.formula = mX2, yname = "y", data = dd, weight = "treated")
psatt
round(summary(psatt)$estimates[1], 4)
round(exp(summary(psatt, type = "RR")$estimates[1]), 3)
round(exp(summary(psatt, type = "OR")$estimates[1]), 3)
```

Compare the results here with those obtained by g-formula in item 4(e)
and with the true contrasts in item 4(f) above.

## Targeted maximum likelihood estimation (TMLE)

We now consider now another double robust approach, known as targeted
maximum likelihood estimation (TMLE). It also corrects the estimator
obtained from the outcome model by elements that are derived from the
exposure model. See [Schuler and Rose
(2017)](https://doi.org/10.1093/aje/kww165) for more details

1.  The first step is to utilize the propensity scores obtained above
    and define the so called clever covariates

```{r clever covariates}
dd$H1 <- dd$x / dd$PS2
dd$H0 <- (1 - dd$x) / (1 - dd$PS2)
```

2.  Then, a working model is fitted for the outcome, in which the clever
    covariates are explanatory variables, but the model also includes
    the previously fitted linear predictor
    $\widehat{\eta}_i = \text{logit}(\widehat Y_i)$ from the original
    outcome model `mY` as an offset term. Moreover, the intercept is
    removed.

```{r model with clever covariates}
epsmod <- glm(y ~ -1 + H0 + H1 + offset(qlogis(yh)),
  family = binomial(link = logit), data = dd
)
eps <- coef(epsmod)
eps
```

3.  The logit-transformed predicted values $\widetilde{Y}_i^{X_i=1}$ and
    $\widetilde{Y}_i^{X_i=0}$ of counterfactual individual risks from
    the original outcome model are now corrected by the estimated
    coefficients of the clever covariates, and the corrected predictions
    are returned to the original scale.

```{r tmle estimates}
yp0.H <- plogis(qlogis(dd$yp0) + eps[1] / (1 - dd$PS2))
yp1.H <- plogis(qlogis(dd$yp1) + eps[2] / dd$PS2)
```

-   

Estimates of the causal contrasts:

```{r tmle-estimates}
EY0.t <- mean(yp0.H)
EY1.t <- mean(yp1.H)
round(Contr(EY1.t, EY0.t), 4)
```

Compare these with previous results and with the true values.

## TMLE with SuperLearner

Let us finally apply some fashionable tools of statistical learning, aka
*machine learning*, using the package `SuperLearner` to fit flexible
models for both exposure and outcome. As this method is computationally
much more demanding, we illustrate its use by a sample of 2000 subjects
only.

1.  A simple random sample of $n=2000$ is drawn from the population.

```{r sample}
set.seed(7622)
n <- 2000
sampind <- sample(N, n)
samp <- dd[sampind, ]
```

2.  The algorithms to be used in this exercise are chosen

```{r algorithms}
SL.library <- c(
  "SL.glm", "SL.step", "SL.step.interaction",
  "SL.glm.interaction", "SL.gam",
  "SL.randomForest", "SL.rpart"
)
```

3.  Function `tmle()` computes estimates of the causal contrasts of
    interest. Argument `A` is for the exposure variable, and argument
    `W` contains the confounders.

The run can take a while ...

```{r tmle SL, eval = FALSE}
tmlest <- tmle(
  Y = samp$y, A = samp$x, W = samp[, c("z1", "z2", "z3", "z4")],
  family = "binomial", Q.SL.library = SL.library,
  g.SL.library = SL.library
)
summary(tmlest)
```

Let us take a closer look at the results. In the beginning are reported
the fractions by which the separate algorithms contribute to the
combined algorithm. After that are given estimates of the causal
contrasts together with their estimated variances and 95% confidence
intervals. The variance of each contrast (on log-scale for RR and OR) is
estimated as the variance of the empirical influence curve divided by
$n$, the number of i.i.d. units of observation. Furthermore, causal risk
differences are estimated also for those factually exposed and
unexposed, respectively.

Note that because this analysis was based on sample data, the estimates
are most probably deviating from the true values because of pure random
error. Therefore it is not possible to assess the magnitude of a
possible bias from a single sample.

**Homework.** When you have more time, try to run `tmle` on as large
sample as is possible and compare its results with previous ones
computed for the whole target population.
