
```{r, include=FALSE}
knitr::opts_chunk$set(keep.source = TRUE, eps = FALSE, results = "hide", prefix.string = "./graph/occoh-caco")
```

# Nested case-control study and case-cohort study: Risk factors of coronary heart disease

In this exercise we shall apply both the nested case-control (NCC) 
design and the case-cohort (CC)  design in sampling
control subjects from a defined cohort or closed study population. 
The case group comprises those cohort members who die from coronary heart disease (CHD) during a $> 20$ years follow-up of the cohort.  
The risk factors of interest are cigarette smoking, systolic blood pressure,  and total cholesterol level.

Our study population is an occupational cohort comprising 1501 men
working in blue-collar jobs in one Nordic country. 
Eligible subjects had no history of coronary heart disease
when recruited to the study in the early 1990s. 
Smoking habits and many other items were inquired at baseline
by a questionnaire, and 
blood pressure was measured by a research nurse, the values being
 written down on the questionnaire. Serum samples were also taken from the cohort members at the same time and were stored in a freezer. For some reason,
the data in the questionnaires were not entered to any computer file, but the questionnaires were kept in a safe storehouse for further purposes. 
Also, no biochemical analyses were initially performed for the sera 
collected from the participants.  However, dates of birth and dates of entry to the study were recorded
in an electronic file.

In 2010 the study was suddenly reactivated by those investigators of the original team who were still alive then.
As the first step mortality follow-up of the cohort members was
executed by record linkage to the national population register, from which
the dates of death and emigration were obtained. Another linkage was performed with the national register of causes of death in order to get the
deaths from coronary heard disease identified. As a result a data file
`occoh.txt` was completed containing the following variables:

| Variable  | Description                                 |
| :-------  | :------------------------------------------ |
| `id` | identification number |
| `birth` | date of birth |
| `entry` | date of recruitment and baseline measurements |
| `exit` | date of exit from mortality follow-up |
| `death` | indicator for vital status at the end of follow-up; 1, if dead from any cause, and = 0, if alive |
|`chdeath` | indicator for death from coronary heart disease; 1, if *yes*, and 0, if *no*. |


This exercise is divided into five main parts:

1. Description of the study base or the follow-up experience of the
  whole cohort, identification of the cases and illustrating the risk sets. 

2. Nested case-control study within the cohort: 
    (i) selection of controls by risk set or time-matched sampling
    using function `ccwc()` in package `Epi`, 
    (ii) collection of exposure data for cases and controls 
    from the pertinent data base of the whole cohort to the
    case-control data set using function `merge()`, and 
    (iii) analysis of the case-control data set with stratified Cox 
    model using function `clogit()` in package `survival()`,

3. Case-cohort study within the cohort:
    (i) selection of a subcohort by simple random sampling from the cohort,
    (ii) collection of exposure data for subcohort members and cases, and
    (iii) analysis of the case-cohort data set with Cox model by weighted partial
    likelihood including appropriate weighting and correction of estimated
    covariance matrix for the model coefficients using function `cch()` in package `survival()`.
 
4. Comparison of results from all previous analyses, also with those from a full cohort design.

5. Further tasks and homework.

  
## Reading the cohort data, illustrating the study base and risk sets

- Load the packages `Epi` and `survival`. 
Read in the cohort data file and name
the resulting data frame as `oc`.
See its structure and print the univariate summaries.
```{r Read in occoh data, echo=TRUE}
library(Epi)
library(survival)
url <- "https://raw.githubusercontent.com/SPE-R/SPE/master/pracs/data"
oc <- read.table(paste(url, "occoh.txt", sep = "/"), header = TRUE)
str(oc)
summary(oc)
```

- It is convenient to change all the dates into fractional calendar years
```{r cal.yr, echo=TRUE}
oc$ybirth <- cal.yr(oc$birth)
oc$yentry <- cal.yr(oc$entry)
oc$yexit <- cal.yr(oc$exit)
```

We shall also compute the age at entry and at exit, respectively,
as age will be the main time scale in our analyses.
```{r age.yr, echo=TRUE}
oc$agentry <- oc$yentry - oc$ybirth
oc$agexit <- oc$yexit - oc$ybirth
```
- As the next step we shall create a `lexis` object
from the data frame along the calendar period and age axes,
and as the outcome event we specify the coronary death.
```{r oclexis, echo=TRUE}
oc.lex <- Lexis(
  entry = list(
    per = yentry,
    age = yentry - ybirth
  ),
  exit = list(per = yexit),
  exit.status = chdeath,
  id = id, data = oc
)
str(oc.lex)
summary(oc.lex)
```

- At this stage it is informative to examine a graphical
presentation of the follow-up lines and outcome cases in a conventional 
Lexis diagram. Make use of the `plot` method for `Lexis` objects. 
Gray lifelines are drawn and a bullet is put at the exit point of those lifelines
that end with the outcome event.
```{r plotlexis, echo=TRUE, fig = FALSE}
par(mfrow = c(1, 1))
plot(oc.lex, xlim = c(1990, 2010), grid = TRUE)
points(oc.lex, pch = c(NA, 16)[oc.lex$lex.Xst + 1])
```

-  As age is here the main time axis, 
we shall graphically illustrate the **study base**, *i.e.* 
the follow-up lines and outcome events, 
only along the age scale, being ordered by age at exit.
<!-- %% Function `plot2()` in the same source file does the work. -->
Vertical lines at those ages when new coronary
deaths occur are drawn to identify the pertinent
**risk sets**. For that purpose it is useful first 
 to sort the data frame and the `Lexis` object 
 jointly by age at exit & age at entry, 
 and to give a new ID number according to that order.
```{r plotlexage, echo=TRUE, fig=FALSE}
oc.ord <- cbind(ID = 1:1501, oc[order(oc$agexit, oc$agentry), ])
oc.lexord <- Lexis(
  entry = list(age = agentry),
  exit = list(age = agexit),
  exit.status = chdeath,
  id = ID, data = oc.ord
)
plot(oc.lexord, "age")
points(oc.lexord, pch = ifelse(oc.lexord$lex.Xst == 1, 16, NA))
with(
  subset(oc.lexord, lex.Xst == 1),
  abline(v = agexit, lty = 3)
)
```
- For a closer look, we now
 zoom the graphical illustration of the risk sets into
event times occurring between 50 to 58 years. -- 
Copy the last four lines from the previous item and add arguments `xlim` and `ylim`
to the call of `plot()`.
```{r plotlexage2, echo=TRUE, fig=FALSE}
plot(oc.lexord, "age", xlim = c(50, 58), ylim = c(5, 65))
points(
  oc.lexord, "age", pch = ifelse(oc.lexord$lex.Xst == 1, 16, NA)
)
with(
  subset(oc.lexord, lex.Xst == 1),
  abline(v = agexit, lty = 3)
)
```

## Nested case-control study

We shall now employ the strategy of **risk-set sampling**
or **time-matched** sampling of controls, *i.e.*
we are conducting a *nested case-control study*
within the cohort.


- The risk sets are defined according to the age at diagnosis of the case. Further matching is applied for age at entry by 1-year agebands.
For this purpose we first generate a categorical variable
`agen2` for age at entry 
```{r agentry2, echo=TRUE}
oc.lex$agen2 <- cut(oc.lex$agentry, br = seq(40, 62, 1))
```

Matched sampling from risk sets may be carried out using 
function `ccwc()` found in the `Epi` package.
Its main arguments are the times 
of `entry` and `exit` which specify the time at risk along the
main time scale (here age), and the outcome variable to be given 
in the `fail` argument. The number of controls per case
is set to be two, and the additional matching factor is given. 

- After setting the RNG seed (with your own number), 
make a call of this function and see
the structure of the resulting data frame `cactrl` 
containing the cases and the chosen individual controls. 
```{r risksetsample, echo=TRUE}
set.seed(98623)
cactrl <-
  ccwc(
    entry = agentry, exit = agexit, fail = chdeath,
    controls = 2, match = agen2,
    include = list(id, agentry),
    data = oc.lex, silent = FALSE
  )
str(cactrl)
```
Check the meaning of the four first columns of the case-control 
data frame from the help page of function `ccwc()`.

- Now we shall start  collecting data on the 
risk factors for the cases and their
matched controls, including determination of the total cholesterol 
levels from the frozen sera! The storehouse of the risk factor measurements for 
the whole cohort is file `occoh-Xdata.txt`. It contains
values of the following variables.

| Variable  | Description                                 |
| :-------  | :------------------------------------------ |
| `id`      | identification number, the same as in `occoh.txt` |
| `smok`    | cigarette smoking with categories; 1: `never`, 2: `former`, 3: `1-14/d`, 4: `15+/d` |
| `sbp`     | systolic blood pressure (mmHg) |
| `tchol`   | total cholesterol level (mmol/l) |

```{r ocX, echo=TRUE}
ocX <- 
  read.table(
    paste(url, "occoh-Xdata.txt", sep = "/"), header = TRUE
  )
str(ocX)
```
 
-  In the next step we collect the values of the risk factors
for our cases and controls by merging the case-control data frame and 
the storehouse file. 
In this  operation we utilize function `merge()` to 
select columns of two data frames: `cactrl` 
(all columns) and `ocX` (four columns) and to merge
these into a single file (see exercise 1.1, subsection 1.1.8, where 
`merge()` was introduced).
 The `id` variable in both files is used as the key to link each
individual case or control with his own data on risk factors.
```{r merge, echo=TRUE}
oc.ncc <- merge(cactrl, ocX[, c("id", "smok", "tchol", "sbp")],
  by = "id"
)
str(oc.ncc)
```

-  We shall treat smoking as categorical and
 total cholesterol and systolic blood pressure 
as quantitative risk factors, but the values of the
latter will be divided by 10 to get more interpretable effect estimates.

| Variable  | Description                                 |
| :-------  | :------------------------------------------ |
| `cholgrp` | cholesterol class; 1: `<5`, 2: `5-<6.5`, 3: `>=6.5` |
| `sbpgrp`  | blood pressure class; 1: `<130`, 2: `130-<150`, 3: `150-<170`, 4: `>=170` |

Convert the smoking variable into a factor.
```{r factor smol, echo=TRUE}
oc.ncc$smok <- factor(oc.ncc$smok,
  labels = c("never", "ex", "1-14/d", ">14/d")
)
```

-  It is useful to start the analysis of case-control data by 
simple tabulations by the categorized risk factors. 
Crude estimates of the rate ratios associated with them,
in which matching is ignored, can be obtained as follows. 
<!-- %% instructed in Janne's lecture -->
<!-- %% on Poisson and logistic models on Saturday 23 May.  -->
We shall focus on smoking
```{r cccrude smok, echo=TRUE}
stat.table(
  index = list(smok, Fail),
  contents = list(count(), percent(smok)),
  margins = TRUE, 
  data = oc.ncc
)
smok.crncc <- glm(Fail ~ smok, family = binomial, data = oc.ncc)
round(ci.exp(smok.crncc), 3)
```



- A proper analysis takes into account matching that was employed 
in the selection of controls for each case from the 
pertinent risk set, further restricted to 
subjects who were about the same age at entry as the case was.
Also, adjustment for the other risk factors is desirable.
In this analysis function `clogit()` in `survival` package is
utilized. It is in fact a wrapper of function `coxph()`. 

```{r clogit , echo=TRUE}
m.clogit <- clogit(Fail ~ smok + I(sbp / 10) + tchol +
  strata(Set), data = oc.ncc)
summary(m.clogit)
round(ci.exp(m.clogit), 3)
```

Compare these with the crude estimates obtained above.

## Case-cohort study

Now we start applying the second major outcome-selective 
sampling strategy
for collecting exposure data from a big study population



-  The subcohort is selected as a  
  simple random sample ($n=260$) from the whole cohort.
The `id`-numbers of the individuals that are 
selected will be stored in vector `subcids`, and
`subcind` is an indicator for inclusion to the subcohort. 
```{r subc sample, echo=TRUE} 
N <- 1501
n <- 260
set.seed(15792)
subcids <- sample(N, n)
oc.lexord$subcind <- 1 * (oc.lexord$id %in% subcids)
```

-  We form the data frame `oc.cc`
 to be used in the subsequent 
analysis selecting the union of the subcohort members 
and the case group from the data frame of the full cohort.
After that we collect the data of the risk factors from the
data storehouse for the  subjects in the case-cohort data 
```{r casecoh data, echo=TRUE} 
oc.cc <- subset(oc.lexord, subcind == 1 | chdeath == 1)
oc.cc <- merge(oc.cc, ocX[, c("id", "smok", "tchol", "sbp")],
  by = "id"
)
str(oc.cc)
```

-  We shall now create a graphical illustration of 
 the lifelines contained in the case-cohort data.
Lines for the subcohort non-cases are grey without bullet at exit,
those for subcohort cases are blue with blue bullet at exit, and
for cases outside the subcohort the lines are red and dotted with
red bullets at exit. 
```{r casecoh-lines, echo=TRUE, fig=FALSE} 
plot(subset(oc.cc, chdeath == 0), "age")
lines(subset(oc.cc, chdeath == 1 & subcind == 1), col = "blue")
lines(subset(oc.cc, chdeath == 1 & subcind == 0), col = "red")
points(subset(oc.cc, chdeath == 1),
  pch = 16,
  col = c("blue", "red")[oc.cc$subcind + 1]
)
```

-  Define the categorical smoking variable again.
```{r grouping , echo=TRUE}
oc.cc$smok <- factor(oc.cc$smok,
  labels = c("never", "ex", "1-14/d", ">14/d")
)
```


<!-- oc.cc$cholgrp <- cut( oc.cc$tchol, br = c(2.4, 5, 6.5, 13),  -->
<!--           include.lowest = T, right = F ) -->
<!-- oc.cc$sbpgrp <- cut( oc.cc$sbp, br = c(95, 130, 150, 170, 240),  -->
<!--           include.lowest = T, right = F ) -->


A crude estimate of the hazard ratio for the various smoking categories $k$
vs. non-smokers ($k=1$) can be obtained by tabulating cases $(D_k)$ and person-years ($y_k$)
 in the subcohort by smoking  and then computing the
relevant exposure odds ratio for each category:
$$ \text{HR}_k ^{\text{crude}} = \frac{D_k/D_1}{y_k/y_1} $$

```{r cc-crude HR by smok}
sm.cc <- stat.table(
  index = smok,
  contents = list(Cases = sum(lex.Xst), Pyrs = sum(lex.dur)),
  margins = TRUE, 
  data = oc.cc
)
print(sm.cc, digits = c(sum = 0, ratio = 1))
HRcc <- 
  (sm.cc[1, -5] / sm.cc[1, 1]) / (sm.cc[2, -5] / sm.cc[2, 1])
round(HRcc, 3)
```
Do these estimates resemble those obtained from nested case-control data?

- To estimate the rate ratios associated with smoking and adjusted for the
other risk factors we now fit the pertinent Cox model 
applying the method of *weighted partial likelihood* as 
presented by Ling & Ying (1993) and Barlow (1994).
This analysis can be done using function `cch()` 
in package `survival` with `method = "LinYing"`

```{r weighted cox LinYing, echo=TRUE} 
oc.cc$survobj <- with(oc.cc, Surv(agentry, agexit, chdeath))
cch.LY <- cch(survobj ~ smok + I(sbp / 10) + tchol,
  stratum = NULL,
  subcoh = ~subcind, id = ~id, cohort.size = N, data = oc.cc,
  method = "LinYing"
)
summary(cch.LY)
```

## Full cohort analysis and comparisons

Finally, suppose the investigators after all could afford to collect the
data on risk factors from the storehouse for the whole cohort.

- Let us form the data frame corresponding to the full cohort design
and convert again smoking to be categorical.

```{r fullcoh, echo=TRUE}
oc.full <- merge(oc.lex, ocX[, c("id", "smok", "tchol", "sbp")],
  by.x = "id", by.y = "id"
)
oc.full$smok <- factor(oc.full$smok,
  labels = c("never", "ex", "1-14/d", ">14/d")
)
```

Juts for comparison with the corresponding analysis in case-cohort data
perform a similar crude estimation of hazard ratios associated with smoking.
```{r cox-crude HR by smok}
sm.coh <- stat.table(
  index = smok,
  contents = list(Cases = sum(lex.Xst), Pyrs = sum(lex.dur)),
  margins = TRUE, 
  data = oc.full
)
print(sm.coh, digits = c(sum = 0, ratio = 1))
HRcoh <- 
  (sm.coh[1, -5] / sm.coh[1, 1]) / (sm.coh[2, -5] / sm.coh[2, 1])
round(HRcoh, 3)
```


- Fit now the ordinary Cox model to the full cohort. There is no need
to employ extra tricks upon the ordinary `coxph()` fit.

```{r cox full, echo=TRUE}
cox.coh <- coxph(Surv(agentry, agexit, chdeath) ~
  smok + I(sbp / 10) + tchol, data = oc.full)
summary(cox.coh)
```

- Lastly, a comparison of the point estimates and standard errors between 
the different designs, including variants of analysis for the case-cohort design, can be performed.

```{r comparison, echo=TRUE}
betas <- cbind(coef(cox.coh), coef(m.clogit), coef(cch.LY))
colnames(betas) <- c("coh", "ncc", "cch.LY")
round(betas, 3)

SEs <- cbind(
  sqrt(diag(cox.coh$var)),
  sqrt(diag(m.clogit$var)),
  sqrt(diag(cch.LY$var))
)
colnames(SEs) <- colnames(betas)
round(SEs, 3)
```

You will notice that the point estimates of the coefficients
obtained from the full cohort, nested case-control, and case-cohort analyses, 
respectively, are somewhat variable.  However,  
the standard errors from the NCC and CC
 analyses should be quite similar when the numbers of cases and non-cases are similar.



## Further exercises and homework

- If you have time, you could run both the NCC study and CC study 
again but now with a larger control group or subcohort;
for example 4 controls per case in NCC and $n=520$ as the subcohort size in CC. 
Remember resetting the seed first. 
Pay attention in the results to how much closer 
will be the point estimates and the proper SEs to those
obtained from the full cohort design. 

-  Instead of simple linear terms for `sbp` and `tchol` you could try to fit
spline models to describe their effects.

-  A popular alternative to weighted partial likelihood
in the analysis of case-cohort data is the *pseudo-likelihood method*
(Prentice 1986), which is based on *late entry* to follow-up 
of the case subjects not belonging to 
the subcohort. 
The way to do this is provided by function `cch()` which
you can apply directly to the case-cohort data `oc.cc` as before
but now with `method = "Prentice"`. -- Try this and compare the results
with those obtained by weighted partial likelihood in model
`cch.LY`. 

- Yet another computational solution for
maximizing weighted partial likelihood is provided by 
a combination of functions `twophase()` 
 and `svycoxph()` of the `survey` package.
<!-- \begin{comment} -->
<!-- In this approach, the starting point is a data frame,  -->
<!-- named e.g. `oc.lex2`, -->
<!-- that contains the whole cohort but includes also  -->
<!-- necessary columns for the interesting risk factors, such that -->
<!-- their values are known for the cases and the subcohort members -->
<!-- but are missing (`NA`) for non-cases outside the subcohort. -->
<!-- The case-cohort design embedded in the cohort  -->
<!-- is viewed as a special case of a *two-phase sampling design*. -->
<!-- The pertinent design features are determined by -->
<!-- the key indicator variables in `oc.lex2`:  -->
<!-- subcohort membership (cf. variable `subcind` above) -->
<!-- and case status (`chdeath`).  -->
<!-- Based on these items,  -->
<!-- a call of function `twophase()` with appropriate -->
<!-- arguments creates an object -->
<!--  of class `twophase`, named e.g. `oc.2ph`, that -->
<!--  contains the original data frame and all the survey design information -->
<!--   (like weights) needed to analyse it. -->
<!-- After that, function `svycoxph()`, which is   -->
<!-- a *survey design version* of `coxph()`,  -->
<!-- is called with a similar model formula as before but in which, -->
<!-- instead of the `data` argument, one would specify `design = oc.2ph`. -->
<!-- % which takes care of the weights.  -->
<!-- \end{comment} -->
The approach is illustrated with an example
 in a vignette *Two-phase designs in epidemiology* by Thomas Lumley
(see <http://cran.r-project.org/web/packages/survey/vignettes/epi.pdf>).
-- You can try this at home and check that you would obtain similar results as
with model `cch.LY`. 




