---
output:
  html_document: default
  pdf_document: default
editor_options:
chunk_output_type: console
---
```{r, include = FALSE}
knitr::opts_chunk$set(keep.source = TRUE, 
                          results = "verbatim", 
                          include = TRUE, 
                             eval = TRUE, 
                          comment = "")
```

```{r echo=FALSE,eval=TRUE}
options(width = 90,
        show.signif.stars = FALSE,
        prompt = " ", continue = " ",
        SweaveHooks = list(fig = function()
                           par(mar = c(3,3,1,1),
            mgp = c(3,1,0) / 1.6,
            las = 1,
            bty = "n", 
           lend = "butt")))
```

# Time-dependent variables and multiple states

The following practical exercise is based on the data from paper:

> P Hovind, L Tarnow, P Rossing, B Carstensen, and HH Parving:
> Improved survival in patients obtaining remission of nephrotic range
>   albuminuria in diabetic nephropathy.
> *Kidney Int*, **66**(3):1180--1186, Sept 2004.

You can find a `.pdf`-version of the paper here:
<http://BendixCarstensen.com/AdvCoh/papers/Hovind.2004.pdf>

## The renal failure dataset

The dataset `renal.dta` contains data on follow up of 125
patients from Steno Diabetes Center. They enter the study when they
are diagnosed with nephrotic range albuminuria (NRA). This is a
condition where the levels of albumin in the urine is exceeds a
certain level as a sign of kidney disease. The levels may however drop
as a consequence of treatment, this is called remission. Patients exit
the study at death or kidney failure (dialysis or transplant).

| Variable  | Description                                 |
| :-------  | :------------------------------------------ |
| `id`      | Patient id |
| `sex`     | 1=male, 2=female |
| `dob`     | Date of birth |
| `doe`     | Date of entry into the study (2.5 years after NRA) |
| `dor`     | Date of remission. Missing if no remission has occurred |
| `dox`     | Date of exit from study |
| `event`   | Exit status: 1,2,3=event (death, ESRD), 0=censored |


1.  The dataset is in Stata-format, so you must read the dataset
    using `read.dta` from the `foreign` package (which is
    part of the standard `R`-distribution). At the same time, convert
    `sex` to a proper factor. Choose where to read the dataset.
    ```{r,  results = "hide" }
    library(Epi)
    library(survival)
    clear()
    library(foreign)
    # renal <- read.dta(
    #  "https://raw.githubusercontent.com/SPE-R/SPE/master/pracs/data/renal.dta")
    renal <- read.dta("http://BendixCarstensen.com/SPE/data/renal.dta")
    renal$sex <- factor(renal$sex, labels = c("M", "F"))
    head(renal)
    ```
2.  Use the `Lexis` function to declare the data as
    survival data with age, calendar time and time since entry into
    the study as timescales. Label any event $>0$ as *ESRD*,
    i.e. renal death (death of kidney (transplant or dialysis), or
    person).
    Note that you must make sure that the *alive* state (here
    `NRA`) is the first, as `Lexis` assumes that
    everyone starts in this state (unless of course
    `entry.status` is specified):
    ```{r }
    Lr <- Lexis(entry = list(per = doe,
                             age = doe - dob,
                             tfi = 0),
                 exit = list(per = dox),
          exit.status = factor(event > 0, labels = c("NRA", "ESRD")),
                 data = renal)
    str(Lr)
    summary(Lr)
    ```
    Make sure you know what the variables in `Lr` stand for.

3.  Visualize the follow-up in a Lexis-diagram, by using the
    `plot` method for `Lexis` objects.
    ```{r Lexis-ups, fig=TRUE}
    plot(Lr, col = "black", lwd = 3)
    subset(Lr, age < 0)
    ```
    What is wrong here? List the data for the person with negative entry age.

4.  Correct the data and make a new plot, for example by:
    ```{r Lexis-def,fig=TRUE}
    Lr <- transform(Lr, age = ifelse(dob > 2000, age + 100, age),
                        dob = ifelse(dob > 2000, dob - 100, dob))
    subset(Lr, id == 586)
    plot(Lr, col = "black", lwd = 3)
    ```

5.  Now make a Cox-regression analysis of ESRD occurrence with
    the variables sex and age at entry into the study, using time
    since entry to the study as time scale.
    ```{r }
    mc <- coxph(Surv(lex.dur, lex.Xst == "ESRD") 
                ~ I(age / 10) + sex, data = Lr)
    summary(mc)
    ```
    What is the hazard ratio between males and females?
    Between two persons who differ 10 years in age at entry?

6.  The main focus of the paper was to assess whether the occurrence of
    remission (return to a lower level of albumin excretion, an
    indication of kidney recovery) influences mortality.
    *Remission* is a time-dependent variable which is initially 0, but
    takes the value 1 when remission occurs. In order to handle this, each
    person who sees a remission must have two records:
  
    -  One record for the time before remission, where entry is
       `doe`, exit is `dor`, remission is 0, and event is 0.
    -  One record for the time after remission, where entry is
       `dor`, exit is `dox`, remission is 1, and event is 0
       or 1 according to whether the person had an event at `dox`.
  
    This is accomplished using the `cutLexis` function on the
    `Lexis` object, where we introduce a remission state `Rem`.
    Also use `split.state=TRUE` to
    have different ESRD states according to whether a person had had
    remission or not prioer to ESRD. The statement to do this is:
    ```{r }
    Lc <- cutLexis(Lr, cut = Lr$dor, # where to cut follow up
                 timescale = "per",  # what timescale are we referring to
                 new.state = "Rem",  # name of the new state
               split.state = TRUE)   # different states depending on previous
    summary(Lc)
    ```
    List the records from a few select persons (choose values for
    `lex.id`, using for example `subset(Lc, lex.id %in% c(5,7,9))`).
    
8.  Now show how the states are connected and the number of transitions
      between them by using `boxes`. This is an interactive command
      that requires you to click in the graph window:
    ```{r eval=FALSE}
    boxes(Lc)
    ```
    It has a couple of fancy arguments, try:
    ```{r Lc-boxes,fig=TRUE}
    boxes(Lc, boxpos = TRUE, scale.R = 100, show.BE = TRUE, hm = 1.5, wm = 1.5)
    ```
    You may even be tempted to read the help page for
    `boxes.Lexis` ...
    
9.  Plot a Lexis diagram where different coloring is
    used for different segments of the follow-up. The
    `plot.Lexis` function draws a line for each record in the
    dataset, so you can index the coloring by `lex.Cst` and
    `lex.Xst` as appropriate --- indexing by a factor corresponds
    to indexing by the *index number* of the factor levels, so you
    must be know which order the factor levels are in:
    ```{r Lexis-rem,fig=TRUE}
    levels(Lc) # names and order of states in lex.Cst and lex.Xst
    par(mai = c(3, 3, 1, 1) / 4, mgp = c(3, 1, 0) / 1.6)
    plot(Lc, col = c("red", "limegreen")[Lc$lex.Cst],
            xlab = "Calendar time", ylab = "Age",
             lwd = 3, grid = 0:20 * 5, las = 1,
            xlim = c(1970, 2010), ylim = c(20, 70), 
            xaxs = "i", yaxs = "i")
    points(Lc, pch = c(NA, NA, 16, 16)[Lc$lex.Xst],
               col = c("red", "limegreen", "transparent", "transparent")[Lc$lex.Cst])
    points(Lc, pch = c(NA, NA, 1, 1)[Lc$lex.Xst],
               col = "black", lwd = 2)
    ```
    
10. Make a Cox-regression of mortality rates (i.e. endpoint `ESRD` or
    `ESRD(Rem)`) with sex, age at entry and remission as
    explanatory variables, using time since entry as timescale, and
    include `lex.Cst` as time-dependent variable, and
    indicate that each record represents follow-up from
    `tfi` to `tfi+lex.dur`. Make sure that you know
    why what goes where here in the call to `coxph`.
    ```{r }
    (EP <- levels(Lc)[3:4])           # define EndPoint states
    m1 <- coxph(Surv(tfi,             # entry time
                     tfi + lex.dur,   # exit time
                     lex.Xst %in% EP) # event
                ~ sex + I((doe - dob - 50) / 10) + # fixed covariates
                  (lex.Cst == "Rem"),              # time-dependent variable
                data = Lc)
    summary(m1)
    ```
    What is the effect of of remission on the rate of ESRD?
    
    ## Splitting the follow-up time
    
    In order to explore the effect of remission on the rate of ESRD, we
    split the data further into small pieces of follow-up. To this
    end we use the function `splitLexis`. The rates can then be
    modeled using a Poisson-model, and the shape of the effect of the 
    underlying *rates* be explored. Furthermore, we can allow effects of both
    time since NRA and current age. To this end we will use splines, so we
    need the `splines` and also the `mgcv` packages.
    
11. Now split the follow-up time every month after entry, and verify
    that the number of events and risk time is the same as before and
    after the split:
    ```{r }
    sLc <- splitLexis(Lc, "tfi", breaks = seq(0, 30, 1/12))
    summary( Lc)
    summary(sLc)
    ```
    
12. Now fit the Poisson-model corresponding to the Cox-model
    we fitted previously. The function `Ns()` produces a model
    matrix corresponding to a piece-wise cubic function, modeling the
    baseline hazard explicitly (think of the `Ns` terms as the
    baseline hazard that is not visible in the Cox-model). You 
    can use the wrapper function `glm.Lexis`
    ```{r }
    mp <- glm.Lexis(sLc, 
                    ~ Ns(tfi, knots = c(0, 2, 5, 10)) +
                      sex + I((doe - dob - 40) / 10) + 
                      I(lex.Cst == "Rem"))
    ci.exp(mp)
    ```
    How does the effects of sex change from the Cox-model?
    
13. Try instead using the `gam` function from the
    `mgcv` package. There is convenience wrapper for this for
    `Lexis` objects as well:
    ```{r }
    library(mgcv)
    mx <- gam.Lexis(sLc,
                    ~ s(tfi, k = 10) + 
                      sex + I((doe - dob - 40) / 10) + 
                      I(lex.Cst == "Rem"))
    ci.exp(mp, subset = c("Cst", "doe", "sex"))
    ci.exp(mx, subset = c("Cst", "doe", "sex"))
    ```
    We see that there is virtually no difference between the two
    approaches in terms of the regression parameters.
    
14. Extract the regression parameters from the models using
      `ci.exp` and compare with the estimates from the Cox-model:
    ```{r }
    ci.exp(mx, subset = c("sex", "dob", "Cst"), pval = TRUE)
    ci.exp(m1)
    round(ci.exp(mp, subset = c("sex", "dob", "Cst")) / ci.exp(m1), 2)
    ```
    How large are the differences in estimated regression parameters?
    
15. The model has the same assumptions as the Cox-model about
    proportionality of rates, but there is an additional assumption that
    the hazard is a smooth function of time since entry. It seems to be
    a sensible assumption (well, restriction) to put on the rates that
    they vary smoothly by time. No such restriction is made in the Cox
    model. The `gam` model optimizes the shape of the smoother by
    general cross-validation. Try to look at the shape of the
    estimated effect of `tfi`:
    ```{r }
    plot(mx)
    ```
    Is this a useful plot?
    
16. However, `plot` (well, `plot.gam`) does not give you the *absolute*
    level of the underlying rates because it bypasses the intercept. So
    in order to predict the rates as a function of `tfi` and the
    covariates, we set up a prediction data frame. Note that age
    in the model specification is entered as `doe-dob`, hence
    the prediction data frame must have these two variables and not
    the age, but it is only the difference that matters for the prediction:
    ```{r pred,fig=TRUE}
    nd <- data.frame(tfi = seq(0, 20, 0.1),
                     sex = "M",
                     doe = 1990,
                     dob = 1940,
                 lex.Cst = "NRA")
    str(nd)
    matshade(nd$tfi, cbind(ci.pred(mp, newdata = nd),
                           ci.pred(mx, newdata = nd)) * 100,
             plot = TRUE,
             type = "l", lwd = 3:4, col = c("black", "forestgreen"),
             log = "y", xlab = "Time since entry (years)",
             ylab = "ESRD rate (per 100 PY) for 50 year old men")
    ```
    Try to overlay with the corresponding prediction from the
    `glm` model using `Ns`.
    
 ## Prediction from the multistate model
    
    If we want to make proper statements about the survival and disease
    probabilities we must know not only how the occurrence of remission
    influences the rate of death/ESRD, but we must also model the
    occurrence rate of remission itself.
    
17. The rates of ESRD were modelled by a Poisson model with
    effects of age and time since NRA --- in the models `mp`
    and `mx`.  But if we want to model whole process we must
    also model the remission rates transition from  `NRA` to
    `Rem`, but the number of events is rather small so we restrict
    covariates in this model to only time since NRA and sex. Note
    that only the records that represent follow-up in the `NRA`
    state should be used; this is most easily done using the
    `gam.Lexis` function
    ```{r rem-inc-mgcv}
    mr <- gam.Lexis(sLc, ~ s(tfi, k = 10) + sex,
                         from = "NRA",
                           to = "Rem")
    summary(mr)
    ci.exp(mr, pval = TRUE)
    ```
    What is the remission rate-ratio between men and women?
    
18. If we want to predict the probability of being in each of the
    three states using these estimated rates, we may resort to
    analytical calculations of the probabilities from the estimated
    rates, which is actually doable in this case, but which will be largely
    intractable for more complicated models.
    Alternatively we can *simulate* the life course for a large
    group of (identical) individuals through a model using the estimated
    rates. That will give a simulated cohort (in the form of a
    `Lexis` object), and we can then just count the number of
    persons in each state at each of a set of time points.
    This is accomplished using the function `simLexis`. The input
    to this is the initial status of the persons whose life-course we
    shall simulate, and the transition rates in suitable form:
    
    - Suppose we want predictions for men aged 50 at
    NRA. The input is in the form of a `Lexis` object (where
    `lex.dur` and `lex.Xst` will be ignored). Note that in
    order to carry over the `time.scales` and the
    `time.since` attributes, we construct the input object using
    `subset` to select columns, and `NULL` to select rows
    (see the example in the help file for `simLexis`):
    ```{r }
    inL <- subset(sLc, select = 1:11)[NULL, ]
    str(inL)
    timeScales(inL)
    inL[1, "lex.id"] <- 1
    inL[1, "per"] <- 2000
    inL[1, "age"] <- 50
    inL[1, "tfi"] <- 0
    inL[1, "lex.Cst"] <- "NRA"
    inL[1, "lex.Xst"] <- NA
    inL[1, "lex.dur"] <- NA
    inL[1, "sex"] <- "M"
    inL[1, "doe"] <- 2000
    inL[1, "dob"] <- 1950
    inL <- rbind(inL, inL)
    inL[2, "sex"] <- "F"
    inL
    str(inL)
    ```
    The other input for the simulation is the models for the transitions. This is given as 
    a list with an element for each transient state (that is `NRA` and
    `Rem`), each of which is again a list with names equal to the
    states that can be reached from the transient state. The content of
    the list will be `glm` objects, in this case the models we
    just fitted, describing the transition rates:
    ```{r }
    Tr <- list("NRA" = list("Rem"  = mr,
                            "ESRD" = mx),
               "Rem" = list("ESRD(Rem)" = mx))
    ```
    With this as input we can now generate a cohort, using `N=5`
    to simulate life course of 10 persons (5 for each set of starting values
    in `inL`):
    ```{r first-sim}
    (iL <- simLexis(Tr, inL, N = 10))
    summary(iL, by = "sex")
    ```
    What type of object have you got as `iL`?.
    Simulate a couple of thousand persons; remember *not* to list the result.
    
19. Now generate the life course of 5,000 persons, and look at the summary.
    The `system.time` command is just to tell you how long it
    took, you may want to start with 1000 just to see how long that takes.
    ```{r 5000-sim}
    system.time(sM <- simLexis(Tr, inL, N = 500, t.range = 12))
    summary(sM, by = "sex")
    ```
    Why are there so many ESRD-events in the resulting data set?
    
20. Now count how many persons are present in each state
    at each time for the first 10 years after entry (which is at age 50). This
    can be done by using `nState`. Try:
    ```{r nState}
    nStm <- nState(subset(sM, sex == "M"), time.scale = "age", 
                   at = seq(0, 10, 0.1), 
                 from = 50)
    nStf <- nState(subset(sM, sex == "F"), time.scale = "age", 
                   at = seq(0, 10, 0.1), 
                 from = 50)
    head(nStf, 15)
    ```
    What is in the object `nStf`?
    
21. With the counts of persons in each state at the
    designated time points (in `nStm`), compute the cumulative
    fraction over the states, arranged in order given by `perm`:
    ```{r pState}
    ppm <- pState(nStm, perm = c(2, 1, 3, 4))
    ppf <- pState(nStf, perm = c(2, 1, 3, 4))
    head(ppf)
    tail(ppf)
    ```
    What do the entries in `ppf` represent?
    
22. Try to plot the cumulative probabilities using the `plot` method
    for `pState` objects: 
    ```{r plot-pp,fig=TRUE}
    plot(ppf)
    ```
    Is this useful?
    
23. Now try to improve the plot so that it is easier to read, and
    easier to compare between men and women, for example:
    ```{r new-nState,fig=TRUE}
    par(mfrow = c(1, 2))
    # Men
    plot(ppm, col = c("limegreen", "red", "#991111", "forestgreen"))
    lines(as.numeric(rownames(ppm)), ppm[, "NRA"], lwd = 2)
    text(59.5, 0.95, "Men", adj = 1, col = "white", font = 2, cex = 1.2)
    axis(side = 4, at = 0:10 / 10)
    axis(side = 4, at = 1:99 / 100, labels = NA, tck = -0.01)
    # Women 
    plot(ppf, col = c("limegreen", "red", "#991111", "forestgreen"),
              xlim = c(60, 50)) # inverted x-axis
    lines(as.numeric(rownames(ppf)), ppf[, "NRA"], lwd = 2)
    text(59.5, 0.95, "Women", adj = 0, col = "white", font = 2, cex = 1.2)
    axis(side = 2, at = 0:10 / 10)
    axis(side = 2, at = 1:99 / 100, labels = NA, tck = -0.01)
    ```
    What is the 10-year risk of remission for men and women respectively?
  