
```{r, include=FALSE}
knitr::opts_chunk$set(results='hide', prefix.string='./graph/tidyverse')
```

# Data manipulation with `tidyverse`

## Introduction

In this chapter we will produce *more or less* the same outputs than in chapter 1.2 and 1.4 using 
`tidyverse` packages framework.

The main objective of this exercise is to get familiar you with some of the main `tidyverse` features.

This is an **optional practical** for participants having already good basic R skills. 
> All the rest of the course can be done without knowledge of `tidyverse`.


## The `births` data

We will work with `births` data-set from `Epi` package. 
First of all, load the `Epi` and `tidyverse` packages. 
Then load the births data-set.

```{r  }
library(Epi)
suppressPackageStartupMessages(library(tidyverse))
data(births) 
```
You can type `?birth` in the `R` console to get a description of the `birth` data-set. (Alternatively, you can refer to chapter  1.3.2)


## `tibble` vs `data.frame`

Most `dplyr` functions outputs return `tibble` object instead of 
`data.frame`.
Inspect the class and characteristics of the `births` object.

```{r  }
class(births)
head(births)
```
**Note:** As any `R` object this can be summarized using `str` function.

```{r  }
str(births)
```
`births` object is a 500 x 8 data.frame. 
Let's convert `births` to `tibble` format with `as_tibble` function.

```{r  }
births_tbl <- as_tibble(births)

class(births_tbl)
births_tbl

# another way to visualize data set is to use glimpse function
glimpse(births_tbl)
```
You can see that `tibble` objects inherits from `data.frame` which implies that all functions 
working with `data.frame` objects will work with `tibble`. The opposite is 
not necessary true. `tibble` has a couple of extra features compared to classical 
`data.frame`. One of them is a slightly more user-friendly console print. The main 
difference is probably that `tibble` objects supports grouping/nesting features. 
Some examples we be done will see latter on.

## Piping functions

This is one of the most popular features of `tidyverse` grammar. It enables function chaining in 
`R`. Function output is transparently passed as input to the next function and so on. 
It can help to make the code more comprehensive and readable.
Here is an example of classic vs piped functions.

```{r  }
head(births, 4)
births |> head(4)
```
**Note:** By default the chained object is given as the first argument to the following 
function.
You can use `.` if this is not the case.  This trick works only with `dplyr` pipe `%>%`

Here is a dummy example where we do not give the first argument to `head` function but the second one.
```{r  }
4 %>% head(births, .)
```

Which could be also be achieve with `base` pipe `|>` with a slightly more complex syntax. 
```{r }
4 |> (\(.) head(births, .))()  # Note the extra parentheses
```

## `mutate` columns

`mutate` will allow you to add and or modify columns in a `tibble`.
Let's create 2 new variables :

  -  `agegrp` (5 years mother's age group) 
  -  `gest4` (gestation time split in 4 categories)

And modify 2 others:

  -  `hyp` (factor version of `hyp`; `normal` vs `hyper`) 
  -  `sex` (factor version of `sex`; `M` vs `F`) 



```{r  }
births_tbl <-
  births_tbl |>
  mutate(
    # modify hyp varible (conversion into factor)
    hyp = factor(hyp, levels = c(0, 1), labels = c("normal", "hyper")),
    # creating a new variable aggrep
    agegrp = cut(matage, breaks = c(20, 25, 30, 35, 40, 45), right = FALSE),
    # modify sex variable (conversion into factor)
    sex = factor(sex, levels = c(1, 2), labels = c("M", "F")),
    # creating a new variable gest4 with case_when instead of cut
    gest4 = 
      case_when(
        gestwks < 25 ~ 'less than 25 weeks',
        gestwks >= 25 & gestwks < 30  ~ '25-30 weeks',
        gestwks >= 30 & gestwks < 35  ~ '30-35 weeks',
        gestwks >= 35   ~ 'more than 35 weeks'
      ) 
  )

births_tbl
```
You can see as header the type of data contained in each column. 
For instance `<dbl>` stands for double (i.e. numeric value) and `fct`
stands for factor. 
In `R` `data.frame` (/ `tibble`) data type
must be the same within a column (e.g. numeric only) but can be of different type across  columns. (note: `matrix` object supports only one type of data) 

Note that `case_when` function do not return a `factor` but a `character` variable in this case.
You will have to force the conversion from `character` to `factor` if needed.

## `select` columns, `filter` and `arrange` rows

`select` is used for column sub-setting while filter is for row sub-setting. 
They are equivalent to the `[]` in `R` base language.
Let's display a table where only babies' `id`, `sex`, `bweight` and mothers' 
`agegrp` are kept for babies with a `bweight` above 4000g.

```{r  }
births_tbl |>
  # select only id, women age group, sex and birth weight of the baby
  select(id, agegrp, sex, bweight) |>
  # keep only babies weighing more than 4000g
  filter(bweight > 4000) 
```
`select` can also be useful to reorder and rename columns.
`arrange` is a nice feature to reorder observations according to chosen attributes.
Let's rename `agegrp`, `sex` and `bweight` with better looking labels (e.g. `Age group`, `Sex`, `Birth weight`) and reorder the table according to babies' decreasing birth weight.

```{r  }
births_tbl |>
  # select only id, women age group, sex and birth weight of the baby
  select(
    id, 
    'Age group' = agegrp, 
    Sex = sex, 
    'Birth weight' = bweight
  ) |>
  # rearrange rows to put the heaviest newborn on top
  arrange(desc(`Birth weight`))
```
**Note:** `tibble` supports blank spaces in the column names which can be handy for final table 
rendering. When you want to work with columns with blank spaces, do not forget to use the $``$
(back-quote).
Try to produce the same table but arranging the rows by decreasing birth weights within each sex.

```{r  }
births_tbl |>
  # select only id, women age group, sex and birth weight of the baby
  select(
    id, 
    'Age group' = agegrp, 
    Sex = sex, 
    'Birth weight' = bweight
  ) |>
  # rearrange rows to put the heaviest newborn on top
  arrange(Sex, desc(`Birth weight`))
```
You can arrange the `tibble` according to more than one column.

## `group_by` and `summarise` data

One greatest features of `dplyr` is the ability to aggregate data sharing a common attribute to process per group operations.
Here we want to compute the number of boys and girls in the data-set.
The idea here is to split the `births` table in two groups. One with the boys, the other with the girls and to count the number of rows in each group.

```{r  }
births.01 <-
  births_tbl |>
  # group the data according to the sex attribute
  group_by(sex) |>
  # count the number of rows/individuals in each group
  summarise(
    count = n()
  )
births.01
```
**Note:** `n` function is equivalent to `nrow` 

Now we have the number of boys and girls, we can compute the distribution (in percentage) of newborns per sex.

```{r  } 
births.02 <-
  births.01 |>
  mutate(
    percent = count / sum(count) * 100
  )
```
**Trick:** most of `dplyr` functions can be combined with a column selection execution statement using
`across` function. This can be very handy in some cases. 
As an example below a code to compute the `sum` of every `birth.02` numerical columns (numerical columns only)

```{r  }
births.03 <-
  births_tbl |>
  select(gest4, sex, gestwks, bweight, matage) |>
  group_by(gest4, sex) |>
  summarise(
    across(
      where(is.numeric),
      ~ mean(.x, na.rm = TRUE)
    ),
    .groups = 'drop'
  )
births.03
```
`across` function supports the purrr-style lambda format, e.g. `~ mean(.x, na.rm = TRUE)`
where `.x` refers to the values from the data set to be passed to the function. This is a common notation you will find across several `tidyverse` functions. 

Some other functions ending by `_with` can be used conditionally within `dplyr`.
As an example we can rename only columns which are not numeric at once (here we want to code all column names using upper characters) using
the combination of `rename_with` and `where`.

```{r  }
births.03 |>
  rename_with(toupper, where(~ !is.numeric(.x)))
```
Let's now compute the number of births and the mean birth weight according to newborn gender.

```{r  }
births.05 <-
  births_tbl |>
  group_by(sex) |>
  summarise(
    count = n(),
    bweight.mean = mean(bweight)
  )
births.05
```
With `births.05` table, compute the global mean birth weight.

**Note:** with such a table the mean baby's birth weight have to be weighted by number of boys and girls (see. `?weighted.mean`). 

```{r  }
births.05 |>
  summarise(
    count.tot = sum(count),
    bweight.mean.tot = weighted.mean(bweight.mean, count)
  )

# this is equivalent to
births_tbl |>
  summarise(
    count.tot = n(),
    bweight.mean.tot = mean(bweight)
  )
```


## Multiple grouping

In some cases, we can be interested in looking at more than a single strata. This can be achieved using multiple grouping.
Let's count the number of births per gender and birth weight class (low vs not low)

```{r  }
births.06 <-
  births_tbl |>
  group_by(sex, lowbw) |>
  summarise(
    count = n()
  )
births.06
```
Try then to compute the percentage of babies in each group.
Look at the difference between the 2 following command lines:

```{r  }
births.06 |>
  mutate(
    percent = count / sum(count) * 100
  )

births.06 |>
  ungroup() |>
  mutate(
    percent = count / sum(count) * 100
  )
```
Are the results the same? 

**Note:** summarizing a data-set will remove the last level of grouping but not the other ones if multiple grouping has been performed. In some cases you might have to explicitly ungroup your `data.frame` before doing further  calculations. 
In the previous examples, if you do not ungroup the data-set, percentages are computed per gender. Ungrouping will let you compute the overall 
percentages. 

**Trick:** a good practice is to always ungroup the summarized dataset in order to prevent form confusion. You can do it using the `.group = 'drop'` option in `summarize()`. 

```{r  }
# this tibble will still be grouped by sex
births_tbl |>
  group_by(sex, lowbw) |>
  summarise(
    count = n()
  )

# this tibble will be group free
births_tbl |>
  group_by(sex, lowbw) |>
  summarise(
    count = n(),
    .groups = 'drop'
  )
```
The same exercise can be done using gestation time group (`gest4`) as stratifying variable.
Lets compute number of births and mean birth weights according to gestation time category.

```{r  }
births_tbl |>
  group_by(gest4) |>
  summarise(
    count = n(),
    bweight.mean = mean(bweight)
  )
```
Any trend?
It seems that birth weight increases with gestation time.
We can also spot that in our data-set the gestation time is missing for 10 newborns. 
We will do not consider this observation for the rest of the exercise.  
Lets cross-tabulate the birth weight category and the gestation time groups.

```{r  }
births_tbl |>
  # keep only the newborn with defined gesational time category
  filter(
    !is.na(gest4)
  ) |>
  group_by(lowbw, gest4) |>
  # compute the number of babies in each cross category
  summarise(
    count = n()
  ) |>
  # compute the percentage of babies in each gestational time category per 
  # birth weight category
  mutate(
    percent = count / sum(count, na.rm = TRUE)
  )
```
Similarly we can be interested in the birth weight distribution per gestational time. 

```{r  }
births_tbl |>
  filter(
    !is.na(gest4)
  ) |>
  group_by(gest4, lowbw) |>
  summarise(
    count = n()
  ) |>
  # compute the percentage of babies in each birth weight category per gestational 
  # time category
  mutate(
    percent = count / sum(count, na.rm = TRUE)
  )
```
**Note:** grouping order matters! and can be confusing so think about ungrouping intermediate tables.

## Bind and join tables

Another nice feature of `dplyr` is tables binding and joining. To practice we will create two `tibble`s:


  -  `age` an individual database which contains `pid` (unique individuals id) 
  and their `age` in year
  -  `center` an study center database which contains `pid` (unique individuals id)
  and `center` (the center where an individual is registered coded as a letter)


```{r  }
age <-
  tibble(
    pid = 1:6,
    age = sample(15:25, size = 6, replace = TRUE)
  )

center <-
  tibble(
    pid = c(1, 2, 3, 4, 10),
    center = c('A', 'B', 'A', 'B', 'C')
  )

age
center
```
Now the tables are define we will try to make the linkage between individuals ages and the center
they belong to.
First of all let's have a look to `bind_rows` function.

```{r  }
bind_rows(age, center)
```
Is it useful?
Here not really because we do not want to *bind* the data-set (but *join* them instead) but that can be in other situations (e.g. several individuals data base to merge..).

**Note:** in `bind_rows`, if columns names do not match, they are fill with `NA`.  

Here we want to join the 2 `tibble` according to their common attribute `pid`. Depending
on the context you can be interested in joining tables differently. Have a look at the differences
between `left_join`, `full_join` and `inner_join`.

```{r  }
# all individuals from ages are kept
left_join(age, center, by = c('pid'))
# everithing is kept
full_join(age, center, by = c('pid'))
# only the individuals present in both dataset are kept
inner_join(age, center, by = c('pid'))
```
Can you spot the differences between the commands above?
As an exercise, you can try to compute the individuals' mean age per center.

```{r  }
inner_join(age, center, by = c('pid')) |>
  group_by(center) |>
  summarise(
    mean_age = mean(age)
  )
```

**Note:** the `by` argument indicates which column should be use to make the *join*. In some 
cases, you might have to uses several columns to match (e.g. per sex and age group), this
can be easily done specifying a vector of column names.

From now on, we will consider other packages than `dplyr` from the `tidyverse` suits.

## Data Visualization with ggplot2

One of the package that have contributed to `tidyverse` success is for sure `ggplot2`.
We will go more into the details on how to produce advanced graphs with `ggplot2` in another practical.
Let's just have a quick example of graphic creation using `ggplot2`.
Let's draw a bar plot to visualize the number of births by women age group.
First you have to create a table with the number of birth per age group.

```{r  }
birth_per_ageg <- births_tbl |> group_by(agegrp) |> summarise(total_births = n())
```

```{r  }
(gg.01 <- 
   ggplot(birth_per_ageg, aes(x = agegrp, y = total_births)) + 
   geom_bar(stat = "identity"))
```
This graph can be customize adding labels and title to the plot:
```{r  }
(gg.02 <- 
   gg.01 +  
   xlab("Women Age Group") + 
   ylab("Total Births") + 
   ggtitle("Number of Births per Women Age Group"))
```
As you can see, plots from `ggplot` family are built incrementally using the `+` operator for each additional element.

## pivoting data with tidyr

`dplyr` often comes with its good friend `tidyr` when we are performing data manipulation. `tidyr` main features is to reshape tables from long to wide format and vis-versa. Let's have an example.
Let's transform in wide format the previously created `birth_per_ageg` table.
We want to have a table with one column per age group containing the `total_births` numbers.

```{r }
birth_per_ageg

birth_per_ageg_wide <- 
  birth_per_ageg |>
  pivot_wider(names_from = 'agegrp', values_from = 'total_births')

birth_per_ageg_wide
```
This table can easily be formatted back in long format using `pivot_longer` function:

```{r }
birth_per_ageg_long <- 
  birth_per_ageg_wide |>
  pivot_longer(cols = 1:5, names_to = 'agegrp', values_to = 'total_births')

birth_per_ageg_long
```
Are the tables `birth_per_ageg` and `birth_per_ageg_long` identical?

```{r }
identical(birth_per_ageg, birth_per_ageg_long)
```
Not really because the factor type of `agegrp` column has been lost during the transformation.
Let's convert `agegrp` column into a factor. Is the new table identical to `birth_per_ageg` ?

```{r }
birth_per_ageg_long_02 <-
  birth_per_ageg_long |>
  mutate(agegrp = as.factor(agegrp))

identical(birth_per_ageg, birth_per_ageg_long_02)
```
Here we have seen the simplest example you can have of table reshaping with `tidyr`. If you are interested check the dedicated vignette (`vignette("pivot")`) to learn how to perform more advanced tables reshaping.

## reading files with readr

Another package from `tidyverse` that can be introduced here is `readr` that contains a set of functions equivalent to the core `R` data.frame reading functions (e.g. `read.table()`, `read.csv()`, `read.delim()`, ...). The main change is that data are loaded in `R` as `tibble` instead of `data.frame`, type of variables (columns) are *guessed* if possible, and some extra data checking tests are performed.

Let's explore this differences with `fem` dataset available in `data` directory.

```{r }
# read a csv using core R
fem.csv.core <- read.csv('data/fem.csv')
# read a csv using tidyverse
fem.csv.tidy <- read_csv('data/fem.csv')
# compare
fem.csv.core
fem.csv.tidy
# table dimensions
dim(fem.csv.core)
dim(fem.csv.tidy)
# compare column types
map(fem.csv.core, class)
map(fem.csv.tidy, class)
```
**note:** in case you do not fully get the last lines and the `map()` call, it will be explained in the next section on `purrr` package. 

Here we see that the only difference is the type of object loaded `data.frame` vs `tibble` and the default type chosen to cast numeric values (`integer` vs `numeric`).

What about loading `occoh.txt` you will be using in some other practical in the coming days.

```{r }
# read a csv using core R
occoh.txt.core <- read.table('data/occoh.txt')
# read a csv using tidyverse
occoh.txt.tidy <- read_table('data/occoh.txt')
occoh.txt.tidy <- read_table('data/occoh.txt')
# compare
occoh.txt.core
occoh.txt.tidy
# table dimensions
dim(occoh.txt.core)
dim(occoh.txt.tidy)
# compare column types
map(occoh.txt.core, class)
map(occoh.txt.tidy, class)
```
As you can see, in addition to inferring the type of columns in the input data (here some dates), using `readr` to load you data-set can help you to detect inconsistencies in input data formatting (there are no true problem here).

If you are interested, you can explore the other functions of `readr` and see how you can tune it.

## String manipulation with stringr

Another popular `tidyverse` popular package is `stringr` package. This package is specialized in the string manipulation. Here are couple of examples.

Let's create a character vector with the following elements representing country names: "Estonia", "Finland", "Denmark", "United Kingdom", "France".
```{r }
countries <- c("Estonia", "Finland", "Denmark", "United Kingdom", "France")
```
With `stringr` functions perform the following actions.

Extract the first three characters from each country name:
```{r }
country_initials <- str_sub(countries, start = 1, end = 3)
```

Convert all country names to uppercase:
```{r }
countries_upper <- str_to_upper(countries)
```

Replace "United" with "Utd" in each country name:
```{r }
countries_modified <- str_replace(countries, "United", "Utd")
```
Find the positions of the letter "n" in each country name:
```{r }
a_positions <- str_locate_all(countries, "n")
```
As you can see, the output of `str_locate_all` is a list (one element per character string) containing a 2 column table with one line for each match. The first column (start) being the position of the beginning of the match and the second one (end) being the end of the match. In our case, since we are searching for a single character match, this 2 indexes are always the same.

Count the number of characters in each country name:
```{r }
character_counts <- str_length(countries)
```
These examples demonstrate various string manipulation operations using the `stringr` package. You can modify the exercises, combine several operations or explore other string manipulation functions provided by `stringr` to further practice and enhance your skills in manipulating and analyzing text data.


## purrr package to apply functions to list

Among my favorite `tidyverse` packages, you will find `purrr`. This package contains
several functions that are very similar to `lapply` function.

Apply a function to each element of the vector using map(). Here producing the mean of some grades per class:

```{r }
# define the grade dataset
grades <- 
  list(
    c1 = c(80, 85, 90), 
    c2 = c(75, 70, 85, 88), 
    c3 = c(90, 85, 95)
  )
# compute grades
mean_grades <- map(grades, mean)
```
By default `map()` return a list. One of the nice feature of `purrr` functions is to be able to specify the type of output you want (e.g. `_dbl` for numeric, `_chr` for characters, ...).
Check and try to explain the differences between the following command lines:

```{r }
map(grades, mean)
map_dbl(grades, mean)
map_chr(grades, mean)
map_df(grades, mean)
```
Other nice features of `map` like functions is he availability to support more than one argument.
`map2()` for 2 arguments and `pmap()` for more than 2. This can be very handy in some conditions.
If you are interested you can have a look to this function help file and play with the examples.

`purrr` package has also a set of functions that can be used to apply iteratively a function using `reduce` and/or `accumulate`. The 2 functions behave the same way, it takes the 2 first element of a list, apply a function taking 2 arguments. The results is combined with the third element of the list and given as input to the same function and so on.. The only difference is that accumulate return intermediate results while reduce return only the final results.

Here an example of the cumulative product of the 10 first numbers.
```{r }
1:10 |> purrr::reduce(`*`)
1:10 |> purrr::accumulate(`*`)
```
`purrr` have many of others useful features. Please check the dedicated documentation if you want to go further with this package.

## Bonus: Rendering tables

Once you have produced a nice data-set we can be interested in rendering it in a nice format that can meet presentation/publication expectations. The `kableExtra` table can be useful to achieve this goal.

```{r  eval=FALSE}
# if(!require(kableExtra)) install.packages('kableExtra')
library(kableExtra)

births.08 <-
  births_tbl |>
  filter(
    !is.na(gest4)
  ) |>
  group_by(gest4) |>
  summarise(
    N = n()
  ) |>
  mutate(
    `(%)` = (N / sum(N)) |> scales::percent()
  )

# default
births.08

# markdown flavor (useful fo automatic report production with knitr)
# births.08 |>
#   knitr::kable(fromat = 'markdown')

# create an html version of the table and save it on the hard drive
births.08 |>
  kable() |>
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE
  ) |>
  save_kable(file = 'births.08.html', self_contained = TRUE)
```

**note:** One other very cool package to produce advance formatted Excel spreadsheet I am using more and more is `openxlsx`. Check it out if you are interested.

```{r echo=FALSE, eval=FALSE}
# trick to create dplyr-s.rnw file.
# this part have to be lauch manually
dplyr_e.path <- '~/OneDrive - IARC/PROJECT/_SPE/SPE/pracs/tidyverse-e.rnw'
dplyr_e <- readLines(dplyr_e.path)
dplyr_s <- purrr::map_chr(dplyr_e, ~ sub('results='hide'', '', .x))
writeLines(dplyr_s, sub('-e.rnw$', '-s.rnw', dplyr_e.path))
```

